### 2016-08-12 template/paper.tex
Initial commit

\documentclass{bmd2016p}

\begin{document}

\begin{center}
\fontsize{14}{20}{\bf Instructions for Preparing a Paper for \\[3pt]
                      the Bicycle and Motorcycle Dynamics Symposium Proceedings}
\end{center}

%%%%%%%%%%%%%%%% authors %%%%%%%%%%%%%%%
\begin{center}
\normalsize{\bf{A. B. Authorone$^{*}$, C. Authortwo$^\#$, 
            D. E Authorthree$^\dag$}}
\end{center}

\begin{center}
\begin{tabular}{c}
$^*$ Faculty of Mechanical Engineering\\
University of Technology\\
Address, Postcode City, Country\\
e-mail: email1@address\\
\end{tabular}
\begin{tabular}{c}
$^\#$ Institute for Mechatronics\\
 University of Technology\\
Address, Postcode City, Country\\
e-mail: email2@address\\
\end{tabular} \\ \vspace{2.5ex}

\begin{tabular}{c}
$^\dag$ Laboratory for Engineering Mechanics\\
 Delft University of Technology\\
 Mekelweg 2, NL-2628~CD~~Delft, The Netherlands\\
 e-mail: D.E.Authorthree@tudelft.nl\\
\end{tabular}
\end{center}

\section*{ABSTRACT}

The first page begins with the title of the paper, the authors, affiliations, 
abstract and keywords as it is shown in this document. The title of the paper 
should be typed in 14\,pt bold font and be centred. It should be separated 
from the symposium name by a 12\,pt space and from authors' names by an 18\,pt 
space. Authors' names should be typed in bold font. Affiliations should be 
typed in a general formatting style, centred within a corresponding bounding 
box. The abstract part of the final paper should not be longer than about 
300~words, so it fits on the title page. The abstract begins with the 
``ABSTRACT'' header typed in bold 11\,pt font. The abstract text should be 
separated from the header by an 11\,pt space. The abstract should be followed 
by a list of keywords after the header ``Keywords:'' typed in bold 11\,pt 
font. The keywords should be separated by commas. A dot should be place after 
the last keyword. Up to about five keywords are allowed. Title, affiliations, 
abstract and keywords have to fit on the title page. The Introduction 
paragraph may start already on the title page provided there is enough space.

\begin{keywords}
guidelines for authors, 
template, 
final symposium paper, 
formatting instructions.
\end{keywords}

\section{INTRODUCTION}

All text should be written in 11\,pt Times New Roman font. General text should 
be justified and divided into logical paragraphs separated by 11\,pt spaces. 
Single line spaces are required. All section headers except abstract and 
references header should be numbered with Arabic numbers without a trailing 
dot. Top level headers should be typed in capital characters. Subsections 
should be numbered with multilevel numbering. A maximum of two sublevels is 
allowed. Subheaders should be typed in normal word case. Font size of 11\,pt 
should be used if not specified otherwise.

\section{GENERAL INSTRUCTIONS}

The paper must be written in English. It must contain the full name, address 
and e-mail address of each author. No hard limits on the total length of the 
paper are set, but it should better not exceed 20 pages. All page margins 
should be 3\,cm. The paper used should have the size 210\,x\,297\, mm 
which is the European (German) A4 size. It is suggested to use styles for 
formatting and automatic reference, figure numbering to avoid editorial 
errors. To avoid compatibility problems it is advised to use only upper or 
lower case Latin alphabet, numbers and the underscore character in the file 
name.

\subsection{Equations}

Equations should be numbered continuously according to the format shown in 
Equation~(\ref{eq:equ1}): 
\begin{equation} \label{eq:equ1}
  e^{i\pi} + 1 = 0, 
\end{equation}
where the unknown symbols are explained after the equation.

\subsection{Figures and tables}

\begin{table}[h!]
\begin{center}
\caption{Example of a table with a short caption.} \label{tab:tab1}
\begin{tabular}{|c|ccc|}
\hline
    &  $x$  &  $y$  &  $z$ \\
\hline
$x'$  &  $\alpha_1$ & $\beta_1$ & $\gamma_1$ \\
$y'$  &  $\alpha_2$ & $\beta_2$ & $\gamma_2$ \\
$z'$  &  $\alpha_3$ & $\beta_3$ & $\gamma_3$ \\
\hline
\end{tabular}
\end{center}
\end{table}
All figures should be clearly readable and relevant to the presented text. Use 
of at least 300\,dpi resolution for pictures and 600\,dpi for line art is 
required, 1\,px wide lines in figures should be avoided as they may become 
invisible in print. There is no limit on the amount of figures as long as they 
do not dominate the text and the total length of the paper is within the 
specified limits. Both figures and tables should be centred on the page.

\begin{figure}[h!]
\begin{center}
  \includegraphics[width=55mm]{figure1}
  \caption{An example of a figure caption. Use 10~pt Times New Roman.
           For long captions, use a text width of 13~cm.
           Use the same style for the tables.} \label{fig:fig1}
\end{center}
\end{figure}
Figures, graphs and tables must be included in the same style as shown for 
Figure~\ref{fig:fig1} and Table~\ref{tab:tab1}.

\subsection{References}

Bibliographical citations should be written in the order in which they are 
cited, see the References section below, where Reference~\cite{Pac02} 
exemplifies the case of a textbook, while Reference~\cite{Ber07} is an article 
in conference proceedings and Reference~\cite{Sha71} is an article in a 
journal. Use can be made of a pre-existing bibtex style which uses a similar 
style.

\section{SUBMISSION OF PAPER}

The paper should be submitted by e-mail to \href{mailto:jbrendelson@bmd2016mke.org}{jbrendelson@bmd2016mke.org} with the email subject containing the phrase \textit{``BMD 2016 paper submission - submitting author last name''}. \uline{The deadline is September 1st, 2016}. The file of the paper must be in the Portable Document Format (PDF) and the file size should be within the limit of 10 MB. Exceptionally, other file formats can be accepted if the symposium organizers have been informed beforehand.

Please make sure that the final version of the paper complies with the standards set here. Please note slightly different formatting requirements of the abstract and the final paper.

\section{CONCLUSIONS}

We very much look forward to welcoming you in Milwaukee! Best wishes and the 
warmest regards from the Organizing Committee of Bicycle and Motorcycle 
Dynamics 2016.

% \bibliographystyle{acm}
% \bibliography{biblio}
\begin{thebibliography}{6}
% \begin{thebibliography}{66} % use this for more than 9 references
\bibitem{Pac02} H.~B.~Pacejka,
\textit{Tyre and Vehicle Dynamics},
Butterworth and Heinemann, Oxford, 2002.

\bibitem{Ber07} E.~Bertolazzi, F.~Biral, M.~Da~Lio and V.~Cossalter,
``The influence of rider's upper body motions on motorcycle minimum time
  maneuvering'',
in C.~L.~Bottasso, P.~Masarati and L.~Trainelli (eds),
\textit{Proceedings, Multibody Dynamics 2007, ECCOMAS Thematic Conference},
Milano, Italy, 25--28 June 2007,
Politecnico di Milano, Milano, 2007, 15~pp.

\bibitem{Sha71} R.~S.~Sharp,
``The stability and control of motorcycles'',
\textit{Proceedings of the IMechE, Part C, Journal of Mechanical Engineering
  Science} \textbf{13} (1971), pp.~316--329.

\end{thebibliography}

\end{document}

### 2016-08-12 latex/paper.tex
Author affiliations

--- 2016-08-12 template/paper.tex
@@
 \begin{center}
-\fontsize{14}{20}{\bf Instructions for Preparing a Paper for \\[3pt]
-                      the Bicycle and Motorcycle Dynamics Symposium Proceedings}
+\fontsize{14}{20}{\bf Buckling and dynamic collapse of the bicycle wheel}
 \end{center}
@@
 \begin{center}
-\normalsize{\bf{A. B. Authorone$^{*}$, C. Authortwo$^\#$, 
-            D. E Authorthree$^\dag$}}
+\normalsize{\bf{M. Ford$^{*}$, J.M. Papadopoulos$^\#$, 
+            O. Balogun$^\dag$}}
 \end{center}
@@
 \begin{tabular}{c}
-$^*$ Faculty of Mechanical Engineering\\
-University of Technology\\
-Address, Postcode City, Country\\
-e-mail: email1@address\\
-\end{tabular}
-\begin{tabular}{c}
-$^\#$ Institute for Mechatronics\\
- University of Technology\\
-Address, Postcode City, Country\\
-e-mail: email2@address\\
-\end{tabular} \\ \vspace{2.5ex}
+$^*$ Department of Mechanical Engineering\\
+Northwestern University\\
+2145 Sheridan Road, Evanston, IL 60208, USA\\
+e-mail: mford@u.northwestern.edu\\[2.5ex]
 
-\begin{tabular}{c}
-$^\dag$ Laboratory for Engineering Mechanics\\
- Delft University of Technology\\
- Mekelweg 2, NL-2628~CD~~Delft, The Netherlands\\
- e-mail: D.E.Authorthree@tudelft.nl\\
+$^\#$ Department of Mechanical and Industrial Engineering\\
+Northeastern University\\
+360 Huntington Ave., Boston, MA 02115, USA\\
+e-mail: j.papadopoulos@neu.edu\\[2.5ex]
+
+$^\dag$ Department of Civil Engineering\\
+Northwestern University\\
+2145 Sheridan Road, Evanston, IL 60208, USA\\
+e-mail: o.balogun@northwestern.edu\\
 \end{tabular}

### 2016-08-12 latex/paper.tex
Add bib entries manually

@@
 \begin{keywords}
-guidelines for authors, 
-template, 
-final symposium paper, 
-formatting instructions.
+bicycle wheel, 
+buckling, 
+bifurcation.
 \end{keywords}
@@
 
-All text should be written in 11\,pt Times New Roman font. General text should 
-be justified and divided into logical paragraphs separated by 11\,pt spaces. 
-Single line spaces are required. All section headers except abstract and 
-references header should be numbered with Arabic numbers without a trailing 
-dot. Top level headers should be typed in capital characters. Subsections 
-should be numbered with multilevel numbering. A maximum of two sublevels is 
-allowed. Subheaders should be typed in normal word case. Font size of 11\,pt 
-should be used if not specified otherwise.
+The bicycle wheel is a prestressed structure and is susceptible to buckling under internal forces.
+As the spokes are tightened uniformly, the rim deforms radially to accommodate the spoke strain.
+At a critical tension, the system reaches a bifurcation point and the rim buckles out of its initial plane.
+The post-buckling configuration is generally stable and the original shape of the wheel can be recovered by reducing the tension.
+Buckling can also be triggered by external forces in an otherwise stable wheel leading to a release of strain energy.
+This is an unstable process which generally leads to catastrophic failure.
+
+Despite its implications for wheel strength, the buckling problem has never received a rigorous treatment to our knowledge.
+Jobst Brandt alludes to buckling in his practical manual for wheelbuilders\cite{Brandt}:
+
+``If the wheel becomes untrue in two large waves during stress relieving, the maximum, safe tension has been exceeded.
+Approach this tension carefully to avoid major rim distortions.
+When the wheel loses alignment from stress relieving, loosen all spokes a half turn before retruing the wheel.''
+
+He did not discuss the problem further, but suggested that wheel failure commonly occurs due to a loss of lateral stability caused by spoke buckling.
+Pippard and Francis\cite{Pip2} derived a model for lateral stiffness based on the elastic foundation model, but did not discuss stability and neglected any effects of spoke tension.
+
+Flexural-torsional buckling of rings can be treated as a special case of buckling of arches, where the included angle is allowed to go to $2\pi$.
+Timoshenko and Gere\cite{Tim} gave a formula for the critical load for a ring with doubly-symmetric cross-section subjected to radial loads.
+The theory of flexural-torsional buckling of monosymmetric arches (bicycle rims have only one plane of symmetry) was broadly formalized by Trahair and Papangelis\cite{Tra} using the virtual work approach to derive the equilibrium and stability equations.
+Their theory has been extended to treat arches restrained by continuous\cite{Brad1} or discrete\cite{Brad2} elastic supports and elastic end restraints\cite{Guo}.
+
+The problem of the prestressed bicycle wheel is unique for a number of reasons.
+First, the buckling load is internal to the structure.
+Second, the spokes act both as elastic restraints resisting buckling and as prestressing elements causing buckling.
+Third, the lateral, radial, and torsional restraining actions of the spokes are coupled---i.e.
+lateral deflection at a spoke may produce lateral, radial, and torsional reactions.
+These considerations extend to other structural systems.
+Large observation wheels such as the London Eye\cite{Mann} and the Singapore Flyer\cite{All} resemble large bicycle wheels and achieve lateral stability due to the bracing angle of prestressed cables, and must be designed against flexural-torsional buckling.
+
+Here we derive a general theory for buckling of a spoked bicycle wheel under self-tension and investigate several special cases.
+Furthermore, we analyze large deformation post-buckling behavior and show how local buckling of individual spokes can lead to collapse of an otherwise sub-critical wheel.
 
@@
 % \begin{thebibliography}{66} % use this for more than 9 references
-\bibitem{Pac02} H.~B.~Pacejka,
-\textit{Tyre and Vehicle Dynamics},
-Butterworth and Heinemann, Oxford, 2002.
+\bibitem{Brandt} J.~Brandt, 
+\textit{The Bicycle Wheel},
+Avocet, Palo Alto, (1993).
 
-\bibitem{Ber07} E.~Bertolazzi, F.~Biral, M.~Da~Lio and V.~Cossalter,
-``The influence of rider's upper body motions on motorcycle minimum time
-  maneuvering'',
-in C.~L.~Bottasso, P.~Masarati and L.~Trainelli (eds),
-\textit{Proceedings, Multibody Dynamics 2007, ECCOMAS Thematic Conference},
-Milano, Italy, 25--28 June 2007,
-Politecnico di Milano, Milano, 2007, 15~pp.
+\bibitem{Pip2} A.J.S.~Pippard, W.E.~Francis, 
+``The Stresses in a Wire Wheel under Side Loads on the Rim'',
+\textit{Philos. Mag.} \textbf{14} (1932), pp. 436-445.
 
-\bibitem{Sha71} R.~S.~Sharp,
-``The stability and control of motorcycles'',
-\textit{Proceedings of the IMechE, Part C, Journal of Mechanical Engineering
-  Science} \textbf{13} (1971), pp.~316--329.
+\bibitem{Tim} S.P.~Timoshenko, J.M.~Gere, 
+\textit{Theory of Elastic Stability}, 
+McGraw-Hill (1961).
+
+\bibitem{Tra} N.S.~Trahair, J.P.~Papangelis, 
+``Flexural-torsional buckling of monosymmetric arches'', 
+\textit{J. Struct. Eng.} \textbf{113} (1987), pp. 2271-2288
+
+\bibitem{Brad1} Y.~-L. Pi and M.~A. Bradford, 
+``Elastic flexural-torsional buckling of continuously restrained arches'',
+\textit{Int. J. Solids Struct.} \textbf{39} (2002), pp. 2299-2322.
+
+\bibitem{Brad2} Y.~-L. Pi and M.~A. Bradford, 
+``Elastic flexural-torsional buckling of discretely restrained arches'',
+\textit{J. Struct. Eng.} \textbf{39} (2002), pp. 2299-2322.
+
+\bibitem{Guo} Y.L.~Guo, S.Y.~Zhao, C.~You, Y.L.~Pi, 
+``Out-of-Plane Elastic Buckling of Circular Arches with Elastic End Restraints'', 
+\textit{J. Struct. Eng.} \textbf{140} (9) (2014)
+
+\bibitem{BandD} C.~J. Burgoyne and R. Dilmaghanian,
+Bicycle Wheel as Prestressed Structure, 
+\textit{J. Eng. Mech.} \textbf{119} (1993), pp. 439-455.
+
+\bibitem{Mann} A.P.~Mann, N.~Thompson, M.~Smits, 
+``Building the British Airways London Eye'', 
+\textit{Proc. Inst. Civ. Eng. - Civ. Eng.} \textbf{144} (2002), pp. 60-72
+
+\bibitem{All} A.~Allsop, P.~Dallard, B.~McNiven, 
+``The Singapore Flyer and Design of Giant Observation Wheels'', 
+\textit{Struct. Eng. Int.-IABSE} \textbf{19}(1), (2009), pp. 12-16
 

### 2016-08-12 latex/paper.tex
Change refs to Bibtex

@@
 Despite its implications for wheel strength, the buckling problem has never received a rigorous treatment to our knowledge.
-Jobst Brandt alludes to buckling in his practical manual for wheelbuilders\cite{Brandt}:
+Jobst Brandt alludes to buckling in his practical manual for wheelbuilders\cite{Brandt1993c}:
 
@@
 He did not discuss the problem further, but suggested that wheel failure commonly occurs due to a loss of lateral stability caused by spoke buckling.
-Pippard and Francis\cite{Pip2} derived a model for lateral stiffness based on the elastic foundation model, but did not discuss stability and neglected any effects of spoke tension.
+Pippard and Francis\cite{Pippard1932d} derived a model for lateral stiffness based on the elastic foundation model, but did not discuss stability and neglected any effects of spoke tension.
 
 Flexural-torsional buckling of rings can be treated as a special case of buckling of arches, where the included angle is allowed to go to $2\pi$.
-Timoshenko and Gere\cite{Tim} gave a formula for the critical load for a ring with doubly-symmetric cross-section subjected to radial loads.
-The theory of flexural-torsional buckling of monosymmetric arches (bicycle rims have only one plane of symmetry) was broadly formalized by Trahair and Papangelis\cite{Tra} using the virtual work approach to derive the equilibrium and stability equations.
-Their theory has been extended to treat arches restrained by continuous\cite{Brad1} or discrete\cite{Brad2} elastic supports and elastic end restraints\cite{Guo}.
+Timoshenko and Gere\cite{Timoshenko1961a} gave a formula for the critical load for a ring with doubly-symmetric cross-section subjected to radial loads.
+The theory of flexural-torsional buckling of monosymmetric arches (bicycle rims have only one plane of symmetry) was broadly formalized by Trahair and Papangelis\cite{Trahair1987b} using the virtual work approach to derive the equilibrium and stability equations.
+Their theory has been extended to treat arches restrained by continuous\cite{Pi2002b} or discrete\cite{Bradford2002d} elastic supports and elastic end restraints\cite{Guo2014b}.
 
@@
 These considerations extend to other structural systems.
-Large observation wheels such as the London Eye\cite{Mann} and the Singapore Flyer\cite{All} resemble large bicycle wheels and achieve lateral stability due to the bracing angle of prestressed cables, and must be designed against flexural-torsional buckling.
+Large observation wheels such as the London Eye\cite{Mann2001a} and the Singapore Flyer\cite{Allsop2009a} resemble large bicycle wheels and achieve lateral stability due to the bracing angle of prestressed cables, and must be designed against flexural-torsional buckling.
 
@@
 
-% \bibliographystyle{acm}
-% \bibliography{biblio}
-\begin{thebibliography}{6}
-% \begin{thebibliography}{66} % use this for more than 9 references
-\bibitem{Brandt} J.~Brandt, 
-\textit{The Bicycle Wheel},
-Avocet, Palo Alto, (1993).
-
-\bibitem{Pip2} A.J.S.~Pippard, W.E.~Francis, 
-``The Stresses in a Wire Wheel under Side Loads on the Rim'',
-\textit{Philos. Mag.} \textbf{14} (1932), pp. 436-445.
-
-\bibitem{Tim} S.P.~Timoshenko, J.M.~Gere, 
-\textit{Theory of Elastic Stability}, 
-McGraw-Hill (1961).
-
-\bibitem{Tra} N.S.~Trahair, J.P.~Papangelis, 
-``Flexural-torsional buckling of monosymmetric arches'', 
-\textit{J. Struct. Eng.} \textbf{113} (1987), pp. 2271-2288
-
-\bibitem{Brad1} Y.~-L. Pi and M.~A. Bradford, 
-``Elastic flexural-torsional buckling of continuously restrained arches'',
-\textit{Int. J. Solids Struct.} \textbf{39} (2002), pp. 2299-2322.
-
-\bibitem{Brad2} Y.~-L. Pi and M.~A. Bradford, 
-``Elastic flexural-torsional buckling of discretely restrained arches'',
-\textit{J. Struct. Eng.} \textbf{39} (2002), pp. 2299-2322.
-
-\bibitem{Guo} Y.L.~Guo, S.Y.~Zhao, C.~You, Y.L.~Pi, 
-``Out-of-Plane Elastic Buckling of Circular Arches with Elastic End Restraints'', 
-\textit{J. Struct. Eng.} \textbf{140} (9) (2014)
-
-\bibitem{BandD} C.~J. Burgoyne and R. Dilmaghanian,
-Bicycle Wheel as Prestressed Structure, 
-\textit{J. Eng. Mech.} \textbf{119} (1993), pp. 439-455.
-
-\bibitem{Mann} A.P.~Mann, N.~Thompson, M.~Smits, 
-``Building the British Airways London Eye'', 
-\textit{Proc. Inst. Civ. Eng. - Civ. Eng.} \textbf{144} (2002), pp. 60-72
-
-\bibitem{All} A.~Allsop, P.~Dallard, B.~McNiven, 
-``The Singapore Flyer and Design of Giant Observation Wheels'', 
-\textit{Struct. Eng. Int.-IABSE} \textbf{19}(1), (2009), pp. 12-16
-
-\end{thebibliography}
+\bibliographystyle{acm}
+\bibliography{C:/Users/matt/OneDrive/Documents/Research/bibtex/Papers-BMD2016.bib}
 

### 2016-08-12 latex/paper.tex
Remove .gitignore files

@@
 2145 Sheridan Road, Evanston, IL 60208, USA\\
-e-mail: o.balogun@northwestern.edu\\
+e-mail: o-balogun@northwestern.edu\\
 \end{tabular}
@@
 
-``If the wheel becomes untrue in two large waves during stress relieving, the maximum, safe tension has been exceeded.
+\begin{quotation}
+\noindent``If the wheel becomes untrue in two large waves during stress relieving, the maximum, safe tension has been exceeded.
 Approach this tension carefully to avoid major rim distortions.
 When the wheel loses alignment from stress relieving, loosen all spokes a half turn before retruing the wheel.''
+\end{quotation}
 
@@
 
-\section{GENERAL INSTRUCTIONS}
+\section{DEFORMATION OF THE WHEEL}
+\subsection{Rim equations}
 
-The paper must be written in English. It must contain the full name, address 
-and e-mail address of each author. No hard limits on the total length of the 
-paper are set, but it should better not exceed 20 pages. All page margins 
-should be 3\,cm. The paper used should have the size 210\,x\,297\, mm 
-which is the European (German) A4 size. It is suggested to use styles for 
-formatting and automatic reference, figure numbering to avoid editorial 
-errors. To avoid compatibility problems it is advised to use only upper or 
-lower case Latin alphabet, numbers and the underscore character in the file 
-name.
+The rim is modeled as an initially circular beam with a constant cross-section.
 
-\subsection{Equations}
+\subsection{Spoke equations}
 
-Equations should be numbered continuously according to the format shown in 
-Equation~(\ref{eq:equ1}): 
-\begin{equation} \label{eq:equ1}
-  e^{i\pi} + 1 = 0, 
+In this paper we consider bicycle wheels with slender prestressed spokes.
+We model a single spoke as an elastic bar pinned at each end.
+This ensures that the spoke force is always collinear with the spoke axis.
+The linearized elongation of a spoke is given by
+\begin{equation}\label{eq:elong}
+\Delta l = (\bm{u}_2 - \bm{u}_1)\cdot \hat{\bm{n}}_0
 \end{equation}
-where the unknown symbols are explained after the equation.
+where $\bm{u}_1$ and $\bm{u}_2$ are the displacement vectors of the two nodes of the spoke and $\hat{\bm{n}}_0$ is a unit vector pointing from node 1 to node 2.
 

### 2016-08-12 latex/paper.tex
Spoke stiffness matrix

@@
 We model a single spoke as an elastic bar pinned at each end.
-This ensures that the spoke force is always collinear with the spoke axis.
+This ensures that the spoke force is always collinear with thes spoke axis.
+We fix our global coordinate system to the hub, which is assumed to be rigid.
+Therefore, the displacement of the hub node of a spoke is zero.
 The linearized elongation of a spoke is given by
-\begin{equation}\label{eq:elong}
-\Delta l = (\bm{u}_2 - \bm{u}_1)\cdot \hat{\bm{n}}_0
+\begin{equation}\label{eq:selong}
+\Delta l = \bm{u}_n\cdot \hat{\bm{n}}_1
 \end{equation}
-where $\bm{u}_1$ and $\bm{u}_2$ are the displacement vectors of the two nodes of the spoke and $\hat{\bm{n}}_0$ is a unit vector pointing from node 1 to node 2.
+where $\bm{u}_n$ is the displacement vector of the spoke nipple and $\hat{\bm{n}}_1$ is a unit vector pointing from the spoke nipple to the hub eyelet. The linearized rotation of the spoke is given by
+\begin{equation}\label{eq:srot1}
+\Omega_1 = \frac{1}{l} \bm{u}_n\cdot \hat{\bm{n}}_2
+\end{equation}
+\begin{equation}\label{eq:srot2}
+\Omega_2 = \frac{1}{l} \bm{u}_n\cdot \hat{\bm{n}}_3
+\end{equation}
+where $\hat{\bm{n}}_2$ and $\hat{\bm{n}}_3$ are two orthogonal unit vectors orthogonal to $\hat{\bm{n}}_1$, and $l$ is the spoke length.
+
+The elongation produces a net force on the rim parallel to the original spoke axis equal to
+\begin{equation}\label{eq:sF1}
+f_1 = \left(\frac{E_sA_s\Delta l}{l}\hat{\bm{n}}_1\right) = \frac{E_sA_s}{l} \hat{\bm{n}}_1 \hat{\bm{n}}_1 \cdot \bm{u}_n 
+\end{equation}
+The rotation of the spoke produces a net force on the rim perpendicular to the original spoke axis due to the rotation of the initial tension in the spoke.
+\begin{equation}\label{eq:sF2}
+f_2 = T \sin{\chi_1} \approx \frac{T}{l} \hat{\bm{n}}_2\hat{\bm{n}}_2 \cdot \bm{u}_n
+\end{equation}
+\begin{equation}\label{eq:sF3}
+f_3 = T \sin{\chi_2} \approx \frac{T}{l} \hat{\bm{n}}_3\hat{\bm{n}}_3 \cdot \bm{u}_n
+\end{equation}
+since $\hat{\bm{n}}_1,\hat{\bm{n}}_2\hat{\bm{n}}_3$ are mutually orthogonal, the total force on the rim is given by $\bm{f} = \bm{k}_f \cdot \bm{u}_n$, where the spoke stiffness matrix is given by
+\begin{equation}\label{eq:kf}
+\bm{k}_f = \frac{E_sA_s}{l}\hat{\bm{n}}_1\hat{\bm{n}}_1 + \frac{T}{l}(\hat{\bm{n}}_2\hat{\bm{n}}_2 + \hat{\bm{n}}_3\hat{\bm{n}}_3)
+\end{equation}
+The first term in Eqn.
+\ref{eq:kf} is the {\bf elastic stiffness} and the second term is the {\bf tension stiffness}.
+The tension stiffness is the component responsible for the transverse vibrations of a guitar string, e.g.
 

### 2016-08-12 latex/paper.tex
Continuum stiffness

@@
 Despite its implications for wheel strength, the buckling problem has never received a rigorous treatment to our knowledge.
-Jobst Brandt alludes to buckling in his practical manual for wheelbuilders\cite{Brandt1993c}:
+Jobst Brandt alludes to buckling in his practical manual for wheelbuilders~\cite{Brandt1993c}:
 
@@
 He did not discuss the problem further, but suggested that wheel failure commonly occurs due to a loss of lateral stability caused by spoke buckling.
-Pippard and Francis\cite{Pippard1932d} derived a model for lateral stiffness based on the elastic foundation model, but did not discuss stability and neglected any effects of spoke tension.
+Pippard and Francis~\cite{Pippard1932d} derived a model for lateral stiffness based on the elastic foundation model, but did not discuss stability and neglected any effects of spoke tension.
 
 Flexural-torsional buckling of rings can be treated as a special case of buckling of arches, where the included angle is allowed to go to $2\pi$.
-Timoshenko and Gere\cite{Timoshenko1961a} gave a formula for the critical load for a ring with doubly-symmetric cross-section subjected to radial loads.
-The theory of flexural-torsional buckling of monosymmetric arches (bicycle rims have only one plane of symmetry) was broadly formalized by Trahair and Papangelis\cite{Trahair1987b} using the virtual work approach to derive the equilibrium and stability equations.
-Their theory has been extended to treat arches restrained by continuous\cite{Pi2002b} or discrete\cite{Bradford2002d} elastic supports and elastic end restraints\cite{Guo2014b}.
+Timoshenko and Gere~\cite{Timoshenko1961a} gave a formula for the critical load for a ring with doubly-symmetric cross-section subjected to radial loads.
+The theory of flexural-torsional buckling of monosymmetric arches (bicycle rims have only one plane of symmetry) was broadly formalized by Trahair and Papangelis~\cite{Trahair1987b} using the virtual work approach to derive the equilibrium and stability equations.
+Their theory has been extended to treat arches restrained by continuous~\cite{Pi2002b} or discrete~\cite{Bradford2002d} elastic supports and elastic end restraints~\cite{Guo2014b}.
 
@@
 These considerations extend to other structural systems.
-Large observation wheels such as the London Eye\cite{Mann2001a} and the Singapore Flyer\cite{Allsop2009a} resemble large bicycle wheels and achieve lateral stability due to the bracing angle of prestressed cables, and must be designed against flexural-torsional buckling.
+Large observation wheels such as the London Eye~\cite{Mann2001a} and the Singapore Flyer~\cite{Allsop2009a} resemble large bicycle wheels and achieve lateral stability due to the bracing angle of prestressed cables, and must be designed against flexural-torsional buckling.
 
@@
 We model a single spoke as an elastic bar pinned at each end.
-This ensures that the spoke force is always collinear with thes spoke axis.
+This ensures that the spoke force is always collinear with the spoke axis.
 We fix our global coordinate system to the hub, which is assumed to be rigid.
@@
 \begin{equation}\label{eq:sF1}
-f_1 = \left(\frac{E_sA_s\Delta l}{l}\hat{\bm{n}}_1\right) = \frac{E_sA_s}{l} \hat{\bm{n}}_1 \hat{\bm{n}}_1 \cdot \bm{u}_n 
+f_1 = \left(\frac{E_sA_s\Delta l}{l}\hat{\bm{n}}_1\right) = \frac{E_sA_s}{l} \hat{\bm{n}}_1 \cdot \bm{u}_n 
 \end{equation}
-The rotation of the spoke produces a net force on the rim perpendicular to the original spoke axis due to the rotation of the initial tension in the spoke.
+The rotation of the spoke produces net forces on the rim perpendicular to the original spoke axis due to the rotation of the initial tension vector.
 \begin{equation}\label{eq:sF2}
-f_2 = T \sin{\chi_1} \approx \frac{T}{l} \hat{\bm{n}}_2\hat{\bm{n}}_2 \cdot \bm{u}_n
+f_2 = T \sin{\Omega_1} \approx \frac{T}{l} \hat{\bm{n}}_2 \cdot \bm{u}_n
 \end{equation}
 \begin{equation}\label{eq:sF3}
-f_3 = T \sin{\chi_2} \approx \frac{T}{l} \hat{\bm{n}}_3\hat{\bm{n}}_3 \cdot \bm{u}_n
+f_3 = T \sin{\Omega_2} \approx \frac{T}{l} \hat{\bm{n}}_3 \cdot \bm{u}_n
 \end{equation}
-since $\hat{\bm{n}}_1,\hat{\bm{n}}_2\hat{\bm{n}}_3$ are mutually orthogonal, the total force on the rim is given by $\bm{f} = \bm{k}_f \cdot \bm{u}_n$, where the spoke stiffness matrix is given by
+since $\hat{\bm{n}}_1,\hat{\bm{n}}_2,\hat{\bm{n}}_3$ are mutually orthogonal, the total force on the rim is given by $\bm{f} = \bm{k}_f \cdot \bm{u}_n$, where the spoke stiffness matrix is given by
 \begin{equation}\label{eq:kf}
@@
 \end{equation}
-The first term in Eqn.
-\ref{eq:kf} is the {\bf elastic stiffness} and the second term is the {\bf tension stiffness}.
+The first term in Eqn.~\ref{eq:kf} is the {\bf elastic stiffness} and the second term is the {\bf tension stiffness}.
 The tension stiffness is the component responsible for the transverse vibrations of a guitar string, e.g.
 
-\subsection{Figures and tables}
-
-\begin{table}[h!]
-\begin{center}
-\caption{Example of a table with a short caption.} \label{tab:tab1}
-\begin{tabular}{|c|ccc|}
-\hline
-    &  $x$  &  $y$  &  $z$ \\
-\hline
-$x'$  &  $\alpha_1$ & $\beta_1$ & $\gamma_1$ \\
-$y'$  &  $\alpha_2$ & $\beta_2$ & $\gamma_2$ \\
-$z'$  &  $\alpha_3$ & $\beta_3$ & $\gamma_3$ \\
-\hline
-\end{tabular}
-\end{center}
-\end{table}
-All figures should be clearly readable and relevant to the presented text. Use 
-of at least 300\,dpi resolution for pictures and 600\,dpi for line art is 
-required, 1\,px wide lines in figures should be avoided as they may become 
-invisible in print. There is no limit on the amount of figures as long as they 
-do not dominate the text and the total length of the paper is within the 
-specified limits. Both figures and tables should be centred on the page.
-
-\begin{figure}[h!]
-\begin{center}
-  \includegraphics[width=55mm]{figure1}
-  \caption{An example of a figure caption. Use 10~pt Times New Roman.
-           For long captions, use a text width of 13~cm.
-           Use the same style for the tables.} \label{fig:fig1}
-\end{center}
-\end{figure}
-Figures, graphs and tables must be included in the same style as shown for 
-Figure~\ref{fig:fig1} and Table~\ref{tab:tab1}.
-
-\subsection{References}
-
-Bibliographical citations should be written in the order in which they are 
-cited, see the References section below, where Reference~\cite{Pac02} 
-exemplifies the case of a textbook, while Reference~\cite{Ber07} is an article 
-in conference proceedings and Reference~\cite{Sha71} is an article in a 
-journal. Use can be made of a pre-existing bibtex style which uses a similar 
-style.
-
-\section{SUBMISSION OF PAPER}
-
-The paper should be submitted by e-mail to \href{mailto:jbrendelson@bmd2016mke.org}{jbrendelson@bmd2016mke.org} with the email subject containing the phrase \textit{``BMD 2016 paper submission - submitting author last name''}.
-\uline{The deadline is September 1st, 2016}.
-The file of the paper must be in the Portable Document Format (PDF) and the file size should be within the limit of 10 MB.
-Exceptionally, other file formats can be accepted if the symposium organizers have been informed beforehand.
-
-Please make sure that the final version of the paper complies with the standards set here. Please note slightly different formatting requirements of the abstract and the final paper.
+Following the approach of Smith~\cite{Smith1901a} and Pippard~\cite{Pippard1932d}, we will approximate the spoke system as a continuum which exerts a distributed load on the rim.
+The stiffness per unit length is defined by the sum of the spoke stiffnesses in one periodic grouping of spokes, divided by the arc length of the grouping.
+\begin{equation}\label{eq:kbar}
+\bar{\bm{k}} = \frac{1}{2\pi R}\left(\frac{n_s}{n_p}\right) \sum_i^{n_p} \bm{k}_{f, i}
+\end{equation}
 
@@
 
-We very much look forward to welcoming you in Milwaukee! Best wishes and the 
-warmest regards from the Organizing Committee of Bicycle and Motorcycle 
-Dynamics 2016.
+...
 
-\bibliographystyle{acm}
+\bibliographystyle{naturemag}
 \bibliography{C:/Users/matt/OneDrive/Documents/Research/bibtex/Papers-BMD2016.bib}

### 2016-08-15 latex/paper.tex
Rim strain energy

@@
 
-The first page begins with the title of the paper, the authors, affiliations, 
-abstract and keywords as it is shown in this document. The title of the paper 
-should be typed in 14\,pt bold font and be centred. It should be separated 
-from the symposium name by a 12\,pt space and from authors' names by an 18\,pt 
-space. Authors' names should be typed in bold font. Affiliations should be 
-typed in a general formatting style, centred within a corresponding bounding 
-box. The abstract part of the final paper should not be longer than about 
-300~words, so it fits on the title page. The abstract begins with the 
-``ABSTRACT'' header typed in bold 11\,pt font. The abstract text should be 
-separated from the header by an 11\,pt space. The abstract should be followed 
-by a list of keywords after the header ``Keywords:'' typed in bold 11\,pt 
-font. The keywords should be separated by commas. A dot should be place after 
-the last keyword. Up to about five keywords are allowed. Title, affiliations, 
-abstract and keywords have to fit on the title page. The Introduction 
-paragraph may start already on the title page provided there is enough space.
+...
 
@@
 The rim is modeled as an initially circular beam with a constant cross-section.
+For simplicity we will assume that the shear center and centroid of the cross-section coincide.
+Except for warping deformation, plane cross-sections remain planar and shear flexibility of the cross-section is neglected.
+Furthermore, we assume that the rim radius is much larger than the height of the rim profile.
+
+The in-plane curvature of the rim is $\kappa_2 = 1/R$ and remains constant during flexural-torsional buckling. The out-of-plane curvature and axial twist rate are
+\begin{equation}\label{eq:k2}
+\kappa_2 = u'' - \frac{\phi}{R}
+\end{equation}
+\begin{equation}\label{eq:k3}
+\kappa_3 = \phi' + \frac{1}{R} u'
+\end{equation}
+The symbol $()'$ indicates a derivative with respect to $s$.
+It can easily be verified that Equations \ref{eq:k2} and \ref{eq:k3} simplify to the standard equations for straight beams in the limit $R\rightarrow \infty$.
+The curvature and twist rate produce a bending moment and twisting moment equal to
+\begin{equation}\label{eq:M2}
+M_2 = EI_{11} \kappa_2
+\end{equation}
+\begin{equation}\label{eq:M3}
+M_3 = GJ \kappa_3 + EI_w \kappa_3''
+\end{equation}
+The strain energy is decomposed into lateral bending, uniform torsion, and warping terms which are related to the curvatures.
+\begin{equation}\label{eq:Urim}
+U_{rim} = \frac{1}{2} EI_{22} \kappa_2^2 + \frac{1}{2} GJ \kappa_3^2 + \frac{1}{2} EI_w (\kappa_3')^2
+\end{equation}
 

### 2016-08-16 latex/paper.tex
Total potential energy

@@
 
-The in-plane curvature of the rim is $\kappa_2 = 1/R$ and remains constant during flexural-torsional buckling. The out-of-plane curvature and axial twist rate are
+The in-plane curvature $\kappa_1$ is assumed to remain constant during flexural-torsional buckling. The in-plane curvature, out-of-plane curvature and twist rate are
+\begin{equation}\label{eq:k1}
+\kappa_1 = 1/R
+\end{equation}
 \begin{equation}\label{eq:k2}
@@
 \end{equation}
-The strain energy is decomposed into lateral bending, uniform torsion, and warping terms which are related to the curvatures.
+The total strain energy is decomposed into lateral bending, uniform torsion, and warping terms which are related to the curvatures.
 \begin{equation}\label{eq:Urim}
-U_{rim} = \frac{1}{2} EI_{22} \kappa_2^2 + \frac{1}{2} GJ \kappa_3^2 + \frac{1}{2} EI_w (\kappa_3')^2
+U_{rim} = \int_0^{2\pi} \left( \frac{1}{2} EI_{22} \kappa_2^2 + \frac{1}{2} GJ \kappa_3^2 + \frac{1}{2} EI_w (\kappa_3')^2 \right)\, ds
+\end{equation}
+After buckling, a differential line segment $ds$ deforms to $dS$. The apparent shortening of the line segment along the $s$ direction is then
+\begin{equation}\label{eq:ds}
+ds - dS = ds - \sqrt{ds^2 + \left(\frac{du}{ds}ds\right)^2} \approx \frac{1}{2} (u')^2 \, ds
+\end{equation}
+The axial compressive force in the rim moves through this differential displacement, performing virtual work equal to
+\begin{equation}\label{eq:Vrimr}
+V_{rim} = \frac{1}{2} \int_0^{2\pi} N_r (u)^2 \, ds
 \end{equation}
@@
 
+\subsection{Total potential energy and buckling criterion}
+The buckled shape of the wheel must be periodic, so we assume a buckled shape of the form
+\begin{equation}\label{eq:modeshape}
+\begin{split}
+u(s) &= \delta u_n \cos{\frac{ns}{R}} \\
+\phi(s) &= \delta\phi_n \cos{\frac{ns}{R}}
+\end{split}
+\end{equation}
+The wheel becomes unstable when the increase in strain energy is exactly balanced by the virtual work for a perturbing displacement $\delta u_n, \delta\phi_n$, i.e. when
+\begin{equation}\label{eq:TotPot}
+U_{rim} - V_{rim} = 0
+\end{equation}
+Substituting the buckling mode shape Eqn.
+\ref{eq:modeshape} into Eqns.
+\ref{eq:Urim} and \ref{eq:Vrimr} and substituting into Eqn.
+\ref{eq:TotPot} yields the critical buckling criterion for the $n$th mode.
+\begin{equation}\label{eq:Tcrit}
+...
+\end{equation}
 \section{CONCLUSIONS}

### 2016-08-16 latex/paper.tex
Stability equation

@@
 \section{DEFORMATION OF THE WHEEL}
+...Tensegrity
+\begin{equation}\label{eq:TN}
+N_r = \frac{n_sT}{2\pi}
+\end{equation}
+
 \subsection{Rim equations}
@@
 The axial compressive force in the rim moves through this differential displacement, performing virtual work equal to
-\begin{equation}\label{eq:Vrimr}
-V_{rim} = \frac{1}{2} \int_0^{2\pi} N_r (u)^2 \, ds
+\begin{equation}\label{eq:Vrim}
+V_{rim} = \frac{1}{2} \int_0^{2\pi} N_r (u')^2 \, ds
 \end{equation}
@@
 \end{equation}
+The distributed load $\bar{\bm{f}}$ does an amount of work per unit length equal to
+\begin{equation}\label{eq:w_spokes}
+\bar{w} = \int_0^{u_n} \bar{\bm{f}} \cdot d\bm{u}_n = \frac{1}{2} \bm{u}_n \cdot \bar{\bm{k}} \cdot \bm{u}_n
+\end{equation}
+The line of action is assumed to pass sufficiently close to the rim shear center that any difference between the displacement of the spoke nipple and the displacement of the shear center can be ignored.
+The total strain energy stored in the spoke system is equal to
+\begin{equation}\label{eq:Us}
+U_{spokes} = \int_0^{2\pi} \bar{w} \, ds = \frac{1}{2}\left( \bar{k}_{uu}u^2 + \bar{k}_{vv}v^2 + \bar{k}_{ww}w^2 + 2\bar{k}_{uv}uv + 2\bar{k}_{uw}uw + 2\bar{k}_{vw}vw\right)
+\end{equation}
 
 \subsection{Total potential energy and buckling criterion}
+
 The buckled shape of the wheel must be periodic, so we assume a buckled shape of the form
@@
 \begin{split}
-u(s) &= \delta u_n \cos{\frac{ns}{R}} \\
-\phi(s) &= \delta\phi_n \cos{\frac{ns}{R}}
+u &= \delta u_n \cos{\frac{ns}{R}} \\
+\phi &= \delta\phi_n \cos{\frac{ns}{R}}
 \end{split}
 \end{equation}
-The wheel becomes unstable when the increase in strain energy is exactly balanced by the virtual work for a perturbing displacement $\delta u_n, \delta\phi_n$, i.e. when
+The wheel becomes unstable when the increase in strain energy is exactly balanced by the virtual work for a perturbing displacement $\delta u_n, \delta\phi_n$, i.e.
+when the total potential energy is zero.
 \begin{equation}\label{eq:TotPot}
-U_{rim} - V_{rim} = 0
+\Pi = U_{rim} + U_{spokes} - V_{rim} = 0
 \end{equation}
@@
 \ref{eq:modeshape} into Eqns.
-\ref{eq:Urim} and \ref{eq:Vrimr} and substituting into Eqn.
-\ref{eq:TotPot} yields the critical buckling criterion for the $n$th mode.
-\begin{equation}\label{eq:Tcrit}
+\ref{eq:Urim}, \ref{eq:Vrim}, and \ref{eq:Us} and substituting into Eqn.
+\ref{eq:TotPot} yields a buckling criterion for the $n$th mode.
+\begin{multline}\label{eq:BuckCrit}
+\frac{\pi EI_{11}}{R^3}(n^2 u_n - R\phi_n)^2 + \frac{\pi n^2}{R^2}\left(GJ + \frac{EI_w}{R^2}n^2\right)(u_n-R\phi_n)^2 + \\
+\pi R(\bar{k}_{uu}^{el} + T\bar{k}_{uu}^{el})u_n^2 - N_r\frac{\pi n^2}{R}u_n^2=0
+\end{multline}
+Equation \ref{eq:BuckCrit} has a quadratic form and can be written using matrix notation:
+\begin{equation}\label{eq:Qform}
+\begin{bmatrix}
+u_n\\\phi_n
+\end{bmatrix} \cdot
+\begin{bmatrix}
+\frac{\partial^2 \Pi}{\partial u_n^2} & \frac{\partial^2 \Pi}{\partial u_n \partial \phi_n}\\
+\frac{\partial^2 \Pi}{\partial u_n\partial\phi_n} & \frac{\partial^2 \Pi}{\partial \phi_n^2}
+\end{bmatrix} \cdot
+\begin{bmatrix}
+u_n\\\phi_n
+\end{bmatrix}
+=0
+\end{equation}
+Non-trivial solutions of Eqn.
+\ref{eq:Qform} exist when the determinant of the matrix vanishes.
+Substituting Eqn.
+\ref{eq:TN} for $N_r$ and solving Eqn.
+\ref{eq:BuckCrit} for $T$ gives the critical spoke tension for the $n$th mode.
+\begin{equation}
 ...
 \end{equation}
+
 \section{CONCLUSIONS}

### 2016-08-17 latex/paper.tex
Indent equations for clarity

@@
 	\begin{equation}\label{eq:sF1}
-f_1 = \left(\frac{E_sA_s\Delta l}{l}\hat{\bm{n}}_1\right) = \frac{E_sA_s}{l} \hat{\bm{n}}_1 \cdot \bm{u}_n 
+	f_1 = E_sA_s\left(\frac{\Delta l}{l}\right)\hat{\bm{n}}_1 = \frac{E_sA_s}{l} \hat{\bm{n}}_1 \cdot \bm{u}_n 
 	\end{equation}
@@
 The line of action is assumed to pass sufficiently close to the rim shear center that any difference between the displacement of the spoke nipple and the displacement of the shear center can be ignored.
-The total strain energy stored in the spoke system is equal to
+Therefore $\bm{u}_n = \bm{u}$.
+Substituting a lateral buckling displacement $\bm{u}=u\hat{\bm{e}}_1$ into Eqn.
+\ref{eq:w_spokes} and integrating over the length of the rim gives the strain energy stored in the spoke system.
 	\begin{equation}\label{eq:Us}
-U_{spokes} = \int_0^{2\pi} \bar{w} \, ds = \frac{1}{2}\left( \bar{k}_{uu}u^2 + \bar{k}_{vv}v^2 + \bar{k}_{ww}w^2 + 2\bar{k}_{uv}uv + 2\bar{k}_{uw}uw + 2\bar{k}_{vw}vw\right)
+	U_{spokes} = \int_0^{2\pi} \bar{w} \, ds = \frac{1}{2} \bar{k}_{uu}u^2
 	\end{equation}

### 2016-08-17 latex/paper.tex
Tc non-dimensional

@@
 	\end{equation}
+The continuum stiffness can be written $\bar{\bm{k}} = \bar{\bm{k}}^{el} + T\bar{\bm{k}}^{tens}$, where the dependence on $T$ is shown explicitly.
+A very close approximation for the $\bar{k}_{uu}$ component for a symmetrically dished wheel is given by
+	\begin{equation}\label{eq:kuu}
+	\bar{k}_{uu} = \frac{n_sE_sA_s}{2\pi Rl}\sin^2{\alpha} + \frac{n_s T}{2\pi Rl}
+	\end{equation}
 
@@
 \ref{eq:TotPot} yields a buckling criterion for the $n$th mode.
-	\begin{multline}\label{eq:BuckCrit}
+	\begin{multline}\label{eq:ECrit}
 	\frac{\pi EI_{11}}{R^3}(n^2 u_n - R\phi_n)^2 + \frac{\pi n^2}{R^2}\left(GJ + \frac{EI_w}{R^2}n^2\right)(u_n-R\phi_n)^2 + \\
@@
 	\end{multline}
-Equation \ref{eq:BuckCrit} has a quadratic form and can be written using matrix notation:
+The torsional rigidity and warping rigidity can be combined into an effective torsional stiffness $\widetilde{GJ} = GJ + n^2EI_w/R^2$.
+Equation \ref{eq:ECrit} has a quadratic form and can be written using matrix notation:
 	\begin{equation}\label{eq:Qform}
@@
 \ref{eq:Qform} exist when the determinant of the matrix vanishes.
-Substituting Eqn.
-\ref{eq:TN} for $N_r$ and solving Eqn.
-\ref{eq:BuckCrit} for $T$ gives the critical spoke tension for the $n$th mode.
-	\begin{equation}
+Expanding the determinant in Eqn.
+\ref{eq:Qform} and substituting Eqn.
+\ref{eq:TN} yields a linear equation for the critical buckling tension of the $n$th mode.
+	\begin{multline}\label{eq:Tc_eq}
+	\left(\frac{2\pi EI_{22}}{R}+\frac{2\pi \widetilde{GJ}n^2}{R} \right) \left( \frac{2\pi \widetilde{GJ}n^2}{R^3} + \frac{2\pi EI_{22}n^4}{R^3} + 2\pi R\bar{k}_{uu}^{el} + \frac{n_sT_{c,n}}{l} - \frac{n_sn^2T_{c,n}}{R}\right)\\
+	- \left( \frac{2\pi EI_{22}n^2}{R^2} + \frac{2\pi \widetilde{GJ}n^2}{R^2} \right)^2 = 0
+	\end{multline}
+Solving Eqn.
+\ref{eq:Tc_eq} for $T_{c,n}$ yields the tension at which the $n$th mode becomes unstable.
+The critical buckling tension $T_c$ is the minimum $T_{c,n}$ with respect to the mode number $n\geq 2$.
+
+\section{BUCKLING DUE TO INTERNAL FORCES}
+The solution to Eqn. \ref{eq:Tc_eq} can be written in non-dimensional form as
+	\begin{equation}\label{eq:Tc_nondim}
+	\frac{T_{c,n}}{T_e} = \left( \frac{\tilde{u}n^2(n^2-1)^2}{1+\tilde{\mu}n^2} + \lambda \right) \left(\frac{1}{n^2-1} \right)
+	\end{equation}
+where the ``Euler tension'' $T_e=2\pi EI_{22}/n_sR^2$ is the tension that would produce a compressive force in the rim equal to the classical buckling load for a straight beam-column of length $2\pi R$.
+The parameter $\tilde{\mu} = \widetilde{GJ}/EI_{22}$ is the ratio of the effective torsional stiffness to the lateral bending stiffness.
+The parameter $\lambda = \bar{k}_{uu}^{el}R^4/EI_{22}$ is the ratio of the spoke system stiffness $\bar{k}_{uu}^{el}R$ to the rim bending stiffness $EI_{22}/R^3$.
+
+\section{SUB-CRITICAL BUCKLING UNDER LATERAL LOADS}
 ...
-	\end{equation}
 
 \section{CONCLUSIONS}
-
 ...

### 2016-08-17 latex/paper.tex
Fit Tc_nondim figure into textwidth
Wheel deformation and schematics

@@
 \documentclass{bmd2016p}
+\usepackage{epstopdf}
 
@@
 \begin{center}
-\normalsize{\bf{M. Ford$^{*}$, J.M. Papadopoulos$^\#$, 
-            O. Balogun$^\dag$}}
+\normalsize{\bf{Matthew Ford$^{*}$, Jim M. Papadopoulos$^\#$, 
+            Oluwaseyi Balogun$^\dag$}}
 \end{center}
@@
 \section{DEFORMATION OF THE WHEEL}
-...Tensegrity
+A typical bicycle wheel (Fig.
+\ref{fig:schematic} consists of a slender rim connected to a hub by means of a system of very slender metal spokes.
+The spokes are usually threaded at the rim side into spoke nipples which are inserted through the rim from the outside.
+The spokes are kept under tension by the rim.
+Without pretension, the spokes would immediately buckle under a very small load due to their high aspect ratio.
+
+\begin{figure}[!h]
+\centering
+\includegraphics[scale=1.0]{figures/bmd_figures-06.eps}
+\caption{caption}
+\label{fig:schematic}
+\end{figure}
+
+To hold the spokes under tension, the rim must be under compression.
+An accurate approximation\footnote{95\% accurate for 6 spokes, $\alpha=10^{\circ}$, $\beta=10^{\circ}$.
+Increasing the number of spokes only improves the estimate.} was given by Sharp~\cite{Sharp1977a} by considering a force balance on the top (or bottom) half of the rim.
 	\begin{equation}\label{eq:TN}
@@
 	\end{equation}
+The spokes act like guy-wires, both supporting the rim radially and stabilizing the rim laterally.
+However, if the spoke tension exceeds a critical value $T_c$, the compression in the rim causes the wheel to buckle laterally, bending and twisting out of its initial plane, as shown in Fig.
+\ref{fig:schematic} (c).
 
@@
 Furthermore, we assume that the rim radius is much larger than the height of the rim profile.
+
+\begin{figure}[!ht]
+\centering
+\includegraphics[scale=1.0]{figures/bmd_figures-07.eps}
+\caption{Deformation of a rim cross-section.
+The shear center $S\rightarrow S'$ translates by the vector $\bm{u} = u\hat{\bm{e}}_1 + v\hat{\bm{e}}_2 + w\hat{\bm{e}}_3$ and rotates through an angle $\phi$.}
+\label{fig:def}
+\end{figure}
 
@@
 
-\section{BUCKLING DUE TO INTERNAL FORCES}
+\section{BUCKLING DUE TO SPOKE TENSION}
 The solution to Eqn. \ref{eq:Tc_eq} can be written in non-dimensional form as
@@
 The parameter $\lambda = \bar{k}_{uu}^{el}R^4/EI_{22}$ is the ratio of the spoke system stiffness $\bar{k}_{uu}^{el}R$ to the rim bending stiffness $EI_{22}/R^3$.
+
+\begin{figure}[!ht]
+\centering
+\includegraphics[scale=1.0]{figures/bmd_figures-05.eps}
+\caption{\textbf{(a)} Buckling mode parameter map.
+Only the first 4 modes are shown.
+\textbf{(b)}--\textbf{(c)} Normalized buckling tension vs.
+$\tilde{\mu}$ and $\lambda$.
+Solid black line = Eqn.
+\ref{eq:Tc_nondim}, dashed red line = power law, blue stars = finite-element simulations.
+In (a) $\lambda=10$ while in (b) $\tilde{\mu} = 0.38$.}
+\label{fig:Tc_nondim}
+\end{figure}
+
+Since $n$ is a discrete variable, there is no closed-form for the minimum $T_{c,n}$ for a given $(\tilde{\mu},\lambda)$.
+However, two important \textit{approximate} solutions can be derived by replacing the discrete $n$ with a continuous variable $\bar{n}$ and minimizing Eqn.
+\ref{eq:Tc_nondim} analytically, resulting in two simple power laws for $T_c$.
+
+\subsection{Rims with low torsional stiffness}\label{sec:powerlaw_1}
+Many bicycle rims are built from thin-walled open channels which have very low torsional stiffness compared with their bending stiffness.
+Replacing the discrete mode number $n$ with a continuous variable $\bar{n}$, we adopt the following ansatz: (a) $\bar{n}^2 \gg 1$ and (b) $\tilde{\mu}\bar{n}^2 \ll 1$.
+Under these conditions, Eqn.
+\ref{eq:Tc_nondim} becomes
+	\begin{equation}\label{eq:Tcn_small_mu}
+	T_{c,n} = \frac{2\pi EI_{11}}{n_sR^2} \left( \tilde{\mu}\bar{n}^4 + \frac{\lambda}{\bar{n}^2}\right)
+	\end{equation}
+Minimizing Eqn. \ref{eq:Tcn_small_mu} with respect to $\bar{n}$ by setting the derivative $dT_{c,n}/d\bar{n}=0$, we obtain the scaling law
+	\begin{equation}\label{eq:nbar_small_mu}
+	\bar{n} = \left(\frac{\lambda}{2\tilde{\mu}} \right)
+	\end{equation}
+Substituting Eqn. \ref{eq:nbar_small_mu} into Eqn. \ref{eq:Tc_small_mu} yields, after simplification, a simple power law relationship.
+	\begin{equation}\label{eq:Tc_small_mu}
+	T_c \approx \frac{11.875}{n_sR^2} \, \left(\bar{k}_{uu}^{el}R^4 \right)^{2/3} \widetilde{GJ}^{1/3}
+	\end{equation}
+Equation \ref{eq:Tc_small_mu} suggests that when the torsional stiffness is smaller than the bending stiffness, the buckling behavior is dominated by torsion and does not depend explicitly on the bending stiffness at all.
+Equation \ref{eq:Tc_small_mu} is shown in Fig.
+\ref{fig:Tc_nondim} next to Equation \ref{eq:Tc_nondim} and finite-element calculations.
+The power law is strictly less than Equation \ref{eq:Tc_nondim} which conveniently gives a conservative design criterion.
+
+\subsection{Spokes much stiffer than the rim}
+A second power law regime appears when $\lambda \gg 1$ and $\tilde{\mu} \sim 1$. Under these conditions, Eqn. \ref{eq:Tc_nondim} becomes
+	\begin{equation}\label{eq:Tcn_boef}
+	T_{c,n} = \frac{2\pi EI_{11}}{n_sR^2}\left(\bar{n}^2 + \frac{\lambda}{\bar{n}^2} \right)
+	\end{equation}
+Following the same procedure in section \ref{sec:powerlaw_1} gives a separate power law, after simplification.
+	\begin{equation}\label{eq:Tc_boef}
+	T_c \approx \frac{4\pi}{n_s} \left(\bar{k}_{uu}^{el}EI_{11} \right)^{1/2} 
+	\end{equation}
+Substituting Eqn.
+\ref{eq:TN} for $T_c$ gives $N_{rc}=2\sqrt{\bar{k}_{uu}^{el}EI_{11}}$, which is precisely the buckling load for an infinite, straight beam on an elastic foundation~\cite{Hetenyi1946a}.
+Figure \ref{fig:Tc_nondim} (c) shows a comparison of Eqn.
+\ref{eq:Tc_boef} against
 

### 2016-08-18 latex/paper.tex
Fix figure order
Add term list

@@
 \usepackage{epstopdf}
+
+\newcommand{\e}{\ensuremath{\hat{\bm{e}}_1}}
+\newcommand{\ee}{\ensuremath{\hat{\bm{e}}_2}}
+\newcommand{\eee}{\ensuremath{\hat{\bm{e}}_3}}
+
+\newcommand{\n}{\ensuremath{\hat{\bm{n}}_1}}
+\newcommand{\nn}{\ensuremath{\hat{\bm{n}}_2}}
+\newcommand{\nnn}{\ensuremath{\hat{\bm{n}}_3}}
 
@@
 \section*{ABSTRACT}
-
-...
 
@@
 \centering
-\includegraphics[scale=1.0]{figures/bmd_figures-06.eps}
+\includegraphics[scale=1.0]{figures/bmd_figures-05.eps}
 \caption{caption}
@@
 \centering
-\includegraphics[scale=1.0]{figures/bmd_figures-07.eps}
-\caption{Deformation of a rim cross-section.
-The shear center $S\rightarrow S'$ translates by the vector $\bm{u} = u\hat{\bm{e}}_1 + v\hat{\bm{e}}_2 + w\hat{\bm{e}}_3$ and rotates through an angle $\phi$.}
+\includegraphics[scale=1.0]{figures/bmd_figures-06.eps}
+\caption{Deformation of a rim cross-section. The shear center $S\rightarrow S'$ translates by the vector $\bm{u} = u\e + v\ee + w\eee$ and rotates through an angle $\phi$.}
 \label{fig:def}
@@
 	\begin{equation}\label{eq:selong}
-	\Delta l = \bm{u}_n\cdot \hat{\bm{n}}_1
+	\Delta l = \bm{u}_n\cdot \n
 	\end{equation}
-where $\bm{u}_n$ is the displacement vector of the spoke nipple and $\hat{\bm{n}}_1$ is a unit vector pointing from the spoke nipple to the hub eyelet. The linearized rotation of the spoke is given by
+where $\bm{u}_n$ is the displacement vector of the spoke nipple and \n{} is a unit vector pointing from the spoke nipple to the hub eyelet. The linearized rotation of the spoke is given by
 	\begin{equation}\label{eq:srot1}
-	\Omega_1 = \frac{1}{l} \bm{u}_n\cdot \hat{\bm{n}}_2
+	\Omega_1 = \frac{1}{l} \bm{u}_n\cdot \nn
 	\end{equation}
 	\begin{equation}\label{eq:srot2}
-	\Omega_2 = \frac{1}{l} \bm{u}_n\cdot \hat{\bm{n}}_3
+	\Omega_2 = \frac{1}{l} \bm{u}_n\cdot \nnn
 	\end{equation}
-where $\hat{\bm{n}}_2$ and $\hat{\bm{n}}_3$ are two orthogonal unit vectors orthogonal to $\hat{\bm{n}}_1$, and $l$ is the spoke length.
+where \nn{} and \nnn{} are two orthogonal unit vectors orthogonal to \n{}, and $l$ is the spoke length.
 
@@
 	\begin{equation}\label{eq:sF1}
-	f_1 = E_sA_s\left(\frac{\Delta l}{l}\right)\hat{\bm{n}}_1 = \frac{E_sA_s}{l} \hat{\bm{n}}_1 \cdot \bm{u}_n 
+	f_1 = E_sA_s\left(\frac{\Delta l}{l}\right)\n = \frac{E_sA_s}{l} \n \cdot \bm{u}_n 
 	\end{equation}
@@
 	\begin{equation}\label{eq:sF2}
-	f_2 = T \sin{\Omega_1} \approx \frac{T}{l} \hat{\bm{n}}_2 \cdot \bm{u}_n
+	f_2 = T \sin{\Omega_1} \approx \frac{T}{l} \nn \cdot \bm{u}_n
 	\end{equation}
 	\begin{equation}\label{eq:sF3}
-	f_3 = T \sin{\Omega_2} \approx \frac{T}{l} \hat{\bm{n}}_3 \cdot \bm{u}_n
+	f_3 = T \sin{\Omega_2} \approx \frac{T}{l} \nn \cdot \bm{u}_n
 	\end{equation}
-since $\hat{\bm{n}}_1,\hat{\bm{n}}_2,\hat{\bm{n}}_3$ are mutually orthogonal, the total force on the rim is given by $\bm{f} = \bm{k}_f \cdot \bm{u}_n$, where the spoke stiffness matrix is given by
+since \n{}, \nn{}, \nnn{} are mutually orthogonal, the total force on the rim is given by $\bm{f} = \bm{k}_f \cdot \bm{u}_n$, where the spoke stiffness matrix is given by
 	\begin{equation}\label{eq:kf}
-	\bm{k}_f = \frac{E_sA_s}{l}\hat{\bm{n}}_1\hat{\bm{n}}_1 + \frac{T}{l}(\hat{\bm{n}}_2\hat{\bm{n}}_2 + \hat{\bm{n}}_3\hat{\bm{n}}_3)
+	\bm{k}_f = \frac{E_sA_s}{l}\n\n + \frac{T}{l}(\nn\nn + \nnn\nnn)
 	\end{equation}
@@
 Therefore $\bm{u}_n = \bm{u}$.
-Substituting a lateral buckling displacement $\bm{u}=u\hat{\bm{e}}_1$ into Eqn.
+Substituting a lateral buckling displacement $\bm{u}=u\e$ into Eqn.
 \ref{eq:w_spokes} and integrating over the length of the rim gives the strain energy stored in the spoke system.
@@
 	\end{equation}
-The continuum stiffness can be written $\bar{\bm{k}} = \bar{\bm{k}}^{el} + T\bar{\bm{k}}^{tens}$, where the dependence on $T$ is shown explicitly.
+The continuum spoke stiffness can be written $\bar{\bm{k}} = \bar{\bm{k}}^{el} + T\bar{\bm{k}}^{tens}$, where the dependence on $T$ is shown explicitly.
 A very close approximation for the $\bar{k}_{uu}$ component for a symmetrically dished wheel is given by
@@
 \centering
-\includegraphics[scale=1.0]{figures/bmd_figures-05.eps}
+\includegraphics[scale=1.0]{figures/bmd_figures-04.eps}
 \caption{\textbf{(a)} Buckling mode parameter map.
@@
 	\begin{equation}\label{eq:Tc_small_mu}
-	T_c \approx \frac{11.875}{n_sR^2} \, \left(\bar{k}_{uu}^{el}R^4 \right)^{2/3} \widetilde{GJ}^{1/3}
+	T_c \approx \frac{11.875}{n_sR^2} \, \left(\bar{k}_{uu}^{el}R^4 \right)^{2/3} (\widetilde{GJ})^{1/3}
 	\end{equation}
@@
 
+\section*{LIST OF TERMS}
+The following notation is used in this paper:
+
+\begin{tabular}{c|p{12cm}}
+\hline
+\textit{\textbf{Term}} & \textit{\textbf{Definition}} \\
+\hline
+$n_s$		& Number of spokes\\
+$E_s, A_s$	& Spoke Young's modulus and cross-sectional area\\
+$l$			& Spoke length\\
+$\alpha$	& Out-of-plane spoke angle--wall angle of a hypothetical ``cone'' of spokes\\
+$\beta$		& In-plane spoke angle--measured between the in-plane projection of the spoke and
+			  a radius of the wheel\\
+$\n$		& Unit vector aligned with the spoke axis\\
+$\nn,\nnn$	& Two unit vectors perpendicular to the spoke axis\\
+$\e,\ee,\eee$ & Coordinate axes defined at the rim shear center\\
+$\bm{u}$	& Displacement of the rim shear center\\
+$u,v,w$		& Vector components of $\bm{u}$ in the \e{}, \ee{}, \eee{} directions.\\
+$\phi$		& Rotation of the rim cross-section about the $\hat{\bm{e}}_3$ axis.\\
+$\delta u, \delta \phi$ & Virtual displacement and rotation\\
+$u_n, \phi_n$ & Buckling mode coefficients\\
+$n$			& Buckling mode number\\
+$T$			& Tension in a spoke\\
+$T_{cn}$	& Critical tension for the $n$th mode\\
+$T_c$		& Critical buckling tension (minimum of $T_{cn}$ over all modes)\\
+$T_e$		& ``Euler'' tension--tension which would produce an axial force in the rim
+			  equal to the buckling load for a straight beam-column of length $2\pi R$\\
+$\bm{f}$	& Force exerted by a spoke on the rim\\
+$f_1, f_2,f_3$ & Vector components of $\bm{f}$\\
+$\bm{k}_f$	& Stiffness matrix for a single spoke\\
+$\bar{\bm{k}}$ & Equivalent stiffness per-unit-length of the spoke system. $\bar{\bm{k}} = \bar{\bm{k}}^{el} + \bar{\bm{k}}^{tens}$\\
+$\bar{\bm{k}}^{el}$ & Part of $\bar{\bm{k}}$ proportional to $E_sA_s$\\
+$\bar{\bm{k}}^{tens}$ & Part of $\bar{\bm{k}}$ proportional to $T$\\
+$\bar{k}_{ij}$ & Tensor components of $\bar{\bm{k}}$\\
+$R$			& Rim radius--measured from the rim shear center\\
+$E$			& Rim Young's modulus\\
+$I_{11}$	& Second moment of area of the rim in the lateral direction\\
+$J$			& Rim torsion constant\\
+$I_w$		& Rim warping constant\\
+$\widetilde{GJ}$ & Effective torsional rigidity including both Saint-Venant torsion and warping\\
+$\tilde{\mu}$ & Ratio of effective torsional rigidity to bending rigidity: $\widetilde{GJ}/EI_{11}$\\
+$\lambda$	& Ratio of spoke system stiffness to rim bending stiffness OR ratio of rim radius to 
+			  the characteristic length of a beam on an elastic foundation.\\
+\hline
+\end{tabular}
+
 \end{document}

### 2016-08-18 latex/paper.tex
finite-element section

@@
 \section*{ABSTRACT}
+A bicycle wheel's spokes must be kept under tension to prevent them from buckling under load. However, above a certain critical tension, the rim will bend and twist out of its plane.
 
@@
 buckling, 
-bifurcation.
+bifurcation,
+rider safety.
 \end{keywords}
@@
 Buckling can also be triggered by external forces in an otherwise stable wheel leading to a release of strain energy.
-This is an unstable process which generally leads to catastrophic failure.
+This is an unstable process which generally leads to a permanently ``taco-ed'' wheel.
 
@@
 
-Flexural-torsional buckling of rings can be treated as a special case of buckling of arches, where the included angle is allowed to go to $2\pi$.
+The bicycle wheel can be modeled as an elastically supported arch in which the included angle is allowed to go to $2\pi$.
 Timoshenko and Gere~\cite{Timoshenko1961a} gave a formula for the critical load for a ring with doubly-symmetric cross-section subjected to radial loads.
@@
 
-Here we derive a general theory for buckling of a spoked bicycle wheel under self-tension and investigate several special cases.
-Furthermore, we analyze large deformation post-buckling behavior and show how local buckling of individual spokes can lead to collapse of an otherwise sub-critical wheel.
+Here we derive a simple solution for buckling of a spoked bicycle wheel under self-tension and investigate several special cases.
 
@@
 \includegraphics[scale=1.0]{figures/bmd_figures-05.eps}
-\caption{caption}
+\caption{\textbf{(a)}-\textbf{(b)} Schematic of a model bicycle wheel.
+In general, the spokes lie on a hyperboloid which forms an angle of $\alpha$ with the wheel plane.
+If the in-plane angle $\beta$ is zero, the spoke surface is precisely a cone.
+\textbf{(c)} A wheel buckled into a saddle shape.}
 \label{fig:schematic}
@@
 	\begin{equation}\label{eq:ds}
-	ds - dS = ds - \sqrt{ds^2 + \left(\frac{du}{ds}ds\right)^2} \approx \frac{1}{2} (u')^2 \, ds
+	ds - dS = ds - \sqrt{ds^2 - \left(\frac{du}{ds}ds\right)^2} \approx \frac{1}{2} (u')^2 \, ds
 	\end{equation}
@@
 Following the approach of Smith~\cite{Smith1901a} and Pippard~\cite{Pippard1932d}, we will approximate the spoke system as a continuum which exerts a distributed load on the rim.
-The stiffness per unit length is defined by the sum of the spoke stiffnesses in one periodic grouping of spokes, divided by the arc length of the grouping.
+The stiffness per unit length is defined by the sum of the spoke stiffnesses in one periodic grouping of $n_p$ spokes, divided by the arc length of the grouping.
 	\begin{equation}\label{eq:kbar}
@@
 	\begin{multline}\label{eq:Tc_eq}
-	\left(\frac{2\pi EI_{22}}{R}+\frac{2\pi \widetilde{GJ}n^2}{R} \right) \left( \frac{2\pi \widetilde{GJ}n^2}{R^3} + \frac{2\pi EI_{22}n^4}{R^3} + 2\pi R\bar{k}_{uu}^{el} + \frac{n_sT_{c,n}}{l} - \frac{n_sn^2T_{c,n}}{R}\right)\\
+	\left(\frac{2\pi EI_{22}}{R}+\frac{2\pi \widetilde{GJ}n^2}{R} \right) \left( \frac{2\pi \widetilde{GJ}n^2}{R^3} + \frac{2\pi EI_{22}n^4}{R^3} + 2\pi R\bar{k}_{uu}^{el} + \frac{n_sT_{cn}}{l} - \frac{n_sn^2T_{cn}}{R}\right)\\
 	- \left( \frac{2\pi EI_{22}n^2}{R^2} + \frac{2\pi \widetilde{GJ}n^2}{R^2} \right)^2 = 0
@@
 Solving Eqn.
-\ref{eq:Tc_eq} for $T_{c,n}$ yields the tension at which the $n$th mode becomes unstable.
-The critical buckling tension $T_c$ is the minimum $T_{c,n}$ with respect to the mode number $n\geq 2$.
+\ref{eq:Tc_eq} for $T_{cn}$ yields the tension at which the $n$th mode becomes unstable.
+The critical buckling tension $T_c$ is the minimum $T_{cn}$ with respect to the mode number $n\geq 2$.
 
@@
 	\begin{equation}\label{eq:Tc_nondim}
-	\frac{T_{c,n}}{T_e} = \left( \frac{\tilde{u}n^2(n^2-1)^2}{1+\tilde{\mu}n^2} + \lambda \right) \left(\frac{1}{n^2-1} \right)
+	\frac{T_{cn}}{T_e} = \left( \frac{\tilde{u}n^2(n^2-1)^2}{1+\tilde{\mu}n^2} + \lambda \right) \left(\frac{1}{n^2-1} \right)
 	\end{equation}
@@
 
-Since $n$ is a discrete variable, there is no closed-form for the minimum $T_{c,n}$ for a given $(\tilde{\mu},\lambda)$.
+Since $n$ is a discrete variable, there is no closed-form for the minimum $T_{cn}$ for a given $(\tilde{\mu},\lambda)$.
 However, two important \textit{approximate} solutions can be derived by replacing the discrete $n$ with a continuous variable $\bar{n}$ and minimizing Eqn.
@@
 	\begin{equation}\label{eq:Tcn_small_mu}
-	T_{c,n} = \frac{2\pi EI_{11}}{n_sR^2} \left( \tilde{\mu}\bar{n}^4 + \frac{\lambda}{\bar{n}^2}\right)
+	T_{cn} = \frac{2\pi EI_{11}}{n_sR^2} \left( \tilde{\mu}\bar{n}^4 + \frac{\lambda}{\bar{n}^2}\right)
 	\end{equation}
-Minimizing Eqn. \ref{eq:Tcn_small_mu} with respect to $\bar{n}$ by setting the derivative $dT_{c,n}/d\bar{n}=0$, we obtain the scaling law
+Minimizing Eqn. \ref{eq:Tcn_small_mu} with respect to $\bar{n}$ by setting the derivative $dT_{cn}/d\bar{n}=0$, we obtain the scaling law
 	\begin{equation}\label{eq:nbar_small_mu}
@@
 	\begin{equation}\label{eq:Tcn_boef}
-	T_{c,n} = \frac{2\pi EI_{11}}{n_sR^2}\left(\bar{n}^2 + \frac{\lambda}{\bar{n}^2} \right)
+	T_{cn} = \frac{2\pi EI_{11}}{n_sR^2}\left(\bar{n}^2 + \frac{\lambda}{\bar{n}^2} \right)
 	\end{equation}
@@
 Figure \ref{fig:Tc_nondim} (c) shows a comparison of Eqn.
-\ref{eq:Tc_boef} against
+\ref{eq:Tc_boef} against Eqn.
+\ref{eq:Tc_nondim} and finite-element calculations.
 
-\section{SUB-CRITICAL BUCKLING UNDER LATERAL LOADS}
-...
+\section{EQUIVALENT SPRING MODEL}
+The lateral stiffness of the wheel can be decomposed into orthogonal modes in the form of Eqn.
+\ref{eq:modeshape}.
+Since the Fourier modes are orthogonal, the total strain energy is the sum of the strain energy in each mode.
+	\begin{equation}\label{eq:Umode}
+	U_{total} = U_0 + U_1 + U_2 + ...
+	\end{equation}
+
+\begin{figure}[!ht]
+\centering
+\includegraphics[scale=1.0]{figures/bmd_figures-03.eps}
+\caption{Equivalent spring model. Each mode stiffness consists of three equivalent springs representing the rim bending and torsion stiffness and the spoke stiffness.}
+\label{fig:Kn}
+\end{figure}
+
+The applied lateral force and torque are determined using Castigliano's first theorem. For an applied force $P$ and no applied torque, we have
+	\begin{equation}\label{eq:Castig}
+	\begin{split}
+	P &= \frac{\partial U_n}{\partial u_n}\\% = \frac{\partial^2 U_n}{\partial u_n^2}u_n + \frac{\partial^2 U_n}{\partial u_n \partial \phi_n}\phi_n\\
+	0 &= \frac{\partial U_n}{\partial \phi_n}% = \frac{\partial^2 U_n}{\partial \phi_n^2}\phi_n + \frac{\partial^2 U_n}{\partial \phi_n \partial u_n}u_n
+	\end{split}
+	\end{equation}
+Substituting $U_{rim} + U_{spokes}$ for $U_n$ and inserting the mode shape Eqn.
+\ref{eq:modeshape}, Eqn.
+\ref{eq:Castig} becomes a linear system of equations for $u_n$ and $\phi_n$.
+Solving \ref{eq:Castig} for $u_n$ yields
+	\begin{equation}
+	u_n = \left(\frac{PR^3}{\pi}\right) \frac{EI_{11}+\widetilde{GJ}n^2}{EI_{11}\widetilde{GJ}n^2(n^2-1)^2 + R\bar{k}_{uu}(EI_{11}+\widetilde{GJ}n^2)}
+	\end{equation}
+The stiffness of the $n$th mode is then defined as
+	\begin{equation}\label{eq:Kn}
+	K_n = P/u_n = \left(\frac{\pi}{R^3}\right) \frac{EI_{11}\widetilde{GJ}n^2(n^2-1)^2}{EI_{11}+\widetilde{GJ}n^2} + \pi R \bar{k}_{uu}
+	\end{equation}
+
+Defining an equivalent torsion stiffness, bending stiffness, and spoke stiffness, Eqn. \ref{eq:Kn} can be written succinctly as
+	\begin{equation}\label{eq:Kn_spring}
+	K_n = \frac{K_{bending}K_{torsion}}{K_{bending}+K_{torsion}} + K_{spokes}
+	\end{equation}
+\,\,\,\, where
+	\begin{equation}\label{eq:Ktorsion}
+	K_{torsion} = \widetilde{GJ}n^2\frac{\pi (n^2-1)^2}{R^3}
+	\end{equation}
+	\begin{equation}\label{eq:Kbending}
+	K_{torsion} = EI_{11}\frac{\pi (n^2-1)^2}{R^3}
+	\end{equation}
+	\begin{equation}\label{eq:Kspokes}
+	K_{spokes} = \pi R \bar{k}_{uu}
+	\end{equation}
+The form of Equation \ref{eq:Kn_spring} suggests an equivalent spring model (Fig.
+\ref{fig:Kn}) for the $n$th mode in which the rim stiffness is represented by two springs connected which is then connected in parallel with the spoke stiffness.
+If there is a large difference between the rim bending stiffness and torsion stiffness, the equivalent rim stiffness will be dominated by the smaller of the two.
+
+With this definition, Eqn. \ref{eq:Tc_nondim} has an extremely simple form
+	\begin{equation}\label{eq:Tc_Kn}
+	T_{cn} = \frac{2R}{n_s(n^2 - R/l)}K_n
+	\end{equation}
+In this form, the benefit of adding spoke stiffness becomes readily apparent. Because the rim stiffness and spoke stiffness act in parallel, increasing either one directly impacts the maximum tension.
+
+% \section{SUB-CRITICAL BUCKLING UNDER LATERAL LOADS}
+% ...
+
+\section{FINITE-ELEMENT MODEL}
+We validated our theoretical predictions using non-linear finite-element calculations implemented in ABAQUS\textsuperscript{\textregistered} 6.14\footnote{ABAQUS\textsuperscript{\textregistered} is a registered trademark of Dassault Syst\`emes.}.
+The spokes and rim were both modeled using 2-node linear beam elements including shear flexibility (element type B31 in the ABAQUS library).
+The rim was interpolated with sufficient elements such that any given beam segment spanned less than 6$^{\circ}$ of the circle.
+
+A controlled eigenstrain was applied to the spokes by defining a coefficient of thermal expansion of 1.0 for the spokes (but not the rim) and ramping the temperature down to 150\% of the expected critical buckling strain.
+Mesh imperfections of the form $\cos{ns/R}$ were added to perturb the initial lateral positions of the rim nodes.
+Modes starting with $n=2$ up to the expected buckling mode were included.
+The ABAQUS Standard static solution procedure was used with non-linear geometric effects.
+
+The average spoke tension was recorded at each time step.
+The buckling tension was determined from a Southwell plot~\cite{Timoshenko1961a} of $u/T$ vs.
+$u$, where $u$ is taken at its maximum location on the rim.
+The deflection of a loaded structure with an initial imperfection is (to first order)
+	\begin{equation}\label{eq:Southwell}
+	u = \frac{a}{T_c/T - 1}
+	\end{equation}
+Therefore the slope of $u/T$ vs.
+$u$ is $1/T_c$.
+The buckling tension was also calculated by finding the strain at which the average tension departs from linearity by more than 3\%.
+The two methods were found to give results in very close agreement.
+
+\begin{figure}[!ht]
+\centering
+\includegraphics[scale=1.0]{figures/bmd_figures-07.eps}
+\caption{\textbf{(a)} Normalized spoke tension vs. lateral deflection. \textbf{(b)} Southwell plot.}
+\label{fig:Southwell}
+\end{figure}
 
@@
 
+\newpage
 \section*{LIST OF TERMS}

### 2016-08-20 latex/paper.tex
Draft abstract and minor edits

@@
 \section*{ABSTRACT}
-A bicycle wheel's spokes must be kept under tension to prevent them from buckling under load. However, above a certain critical tension, the rim will bend and twist out of its plane.
+The spokes of a bicycle wheel must be kept under tension to prevent them from buckling under external loads.
+However, above a certain critical tension, the rim will bend and twist out of its plane into a ``pringle'' or ``taco'' shape.
+Despite its importance for wheel strength and safety, there is no formula in the literature for the maximum tension a wheel can withstand before buckling.
+
+Here we derive such a formula.
+The non-dimensionalized buckling tension depends only on two parameters which are identified as the ratio of torsional stiffness to bending stiffness, and the ratio of spoke stiffness to bending stiffness.
+The number of ``waves'' in the buckled shape is determined by two stiffness ratios and may in general by greater than 2.
+Finally, we show that the wheel may be represented by a system of equivalent springs, two series-connected springs for the rim bending and torsion stiffness, and a parallel-connected spring for the spoke system.
 
@@
 Solving \ref{eq:Castig} for $u_n$ yields
-	\begin{equation}
+	\begin{equation}\label{eq:un}
 	u_n = \left(\frac{PR^3}{\pi}\right) \frac{EI_{11}+\widetilde{GJ}n^2}{EI_{11}\widetilde{GJ}n^2(n^2-1)^2 + R\bar{k}_{uu}(EI_{11}+\widetilde{GJ}n^2)}
 	\end{equation}
-The stiffness of the $n$th mode is then defined as
-	\begin{equation}\label{eq:Kn}
-	K_n = P/u_n = \left(\frac{\pi}{R^3}\right) \frac{EI_{11}\widetilde{GJ}n^2(n^2-1)^2}{EI_{11}+\widetilde{GJ}n^2} + \pi R \bar{k}_{uu}
-	\end{equation}
+% The stiffness of the $n$th mode is then defined as
+	% \begin{equation}\label{eq:Kn}
+	% K_n = P/u_n = \left(\frac{\pi}{R^3}\right) \frac{EI_{11}\widetilde{GJ}n^2(n^2-1)^2}{EI_{11}+\widetilde{GJ}n^2} + \pi R \bar{k}_{uu}
+	% \end{equation}
 
-Defining an equivalent torsion stiffness, bending stiffness, and spoke stiffness, Eqn. \ref{eq:Kn} can be written succinctly as
+Defining an equivalent torsion stiffness, bending stiffness, and spoke stiffness, the mode stiffness $P/u_n$ can be obtained from Eqn. \ref{eq:un} as
 	\begin{equation}\label{eq:Kn_spring}
@@
 	\begin{equation}\label{eq:Tc_Kn}
-	T_{cn} = \frac{2R}{n_s(n^2 - R/l)}K_n
+	T_{cn} = \frac{2RK_n}{n_s(n^2 - R/l)}
 	\end{equation}
@@
 
-\section{FINITE-ELEMENT MODEL}
-We validated our theoretical predictions using non-linear finite-element calculations implemented in ABAQUS\textsuperscript{\textregistered} 6.14\footnote{ABAQUS\textsuperscript{\textregistered} is a registered trademark of Dassault Syst\`emes.}.
+\section{FINITE-ELEMENT CALCULATIONS}
+We validated our theoretical predictions against non-linear finite-element calculations implemented in ABAQUS\textsuperscript{\textregistered} 6.14\footnote{ABAQUS\textsuperscript{\textregistered} is a registered trademark of Dassault Syst\`emes.}.
 The spokes and rim were both modeled using 2-node linear beam elements including shear flexibility (element type B31 in the ABAQUS library).
@@
 
-A controlled eigenstrain was applied to the spokes by defining a coefficient of thermal expansion of 1.0 for the spokes (but not the rim) and ramping the temperature down to 150\% of the expected critical buckling strain.
+A controlled tensioning strain was applied to the spokes by defining a coefficient of thermal expansion of 1.0 for the spokes (but not the rim) and ramping the temperature down to 150\% of the expected critical buckling strain.
 Mesh imperfections of the form $\cos{ns/R}$ were added to perturb the initial lateral positions of the rim nodes.
 Modes starting with $n=2$ up to the expected buckling mode were included.
-The ABAQUS Standard static solution procedure was used with non-linear geometric effects.
+The solution was obtained via an implicit integration procedure (ABAQUS Standard) including non-linear geometric effects.
 
-The average spoke tension was recorded at each time step.
+\begin{figure}[!ht]
+\centering
+\includegraphics[scale=1.0]{figures/bmd_figures-07.eps}
+\caption{\textbf{(a)} Normalized spoke tension vs. lateral deflection. \textbf{(b)} Southwell plot for a simulated wheel with an imperfection.}
+\label{fig:Southwell}
+\end{figure}
+
+The average spoke tension was recorded at each loading step.
 The buckling tension was determined from a Southwell plot~\cite{Timoshenko1961a} of $u/T$ vs.
 $u$, where $u$ is taken at its maximum location on the rim.
-The deflection of a loaded structure with an initial imperfection is (to first order)
+The deflection of a loaded structure with an initial imperfection is approximately
 	\begin{equation}\label{eq:Southwell}
@@
 The two methods were found to give results in very close agreement.
-
-\begin{figure}[!ht]
-\centering
-\includegraphics[scale=1.0]{figures/bmd_figures-07.eps}
-\caption{\textbf{(a)} Normalized spoke tension vs. lateral deflection. \textbf{(b)} Southwell plot.}
-\label{fig:Southwell}
-\end{figure}
 

### 2016-08-20 latex/paper.tex
Simplified K_rim expresion
Minor edits to sections 0-2

@@
 The non-dimensionalized buckling tension depends only on two parameters which are identified as the ratio of torsional stiffness to bending stiffness, and the ratio of spoke stiffness to bending stiffness.
-The number of ``waves'' in the buckled shape is determined by two stiffness ratios and may in general by greater than 2.
-Finally, we show that the wheel may be represented by a system of equivalent springs, two series-connected springs for the rim bending and torsion stiffness, and a parallel-connected spring for the spoke system.
+The number of ``waves'' in the buckled shape is determined by these two stiffness ratios and may in general by greater than 2.
+Finally, we show that the wheel may be represented by a system of equivalent springs: two series-connected springs for the rim bending and torsion stiffness, connected in parallel with a spring for the spoke system.
 
@@
 bifurcation,
-rider safety.
+bicycle safety.
 \end{keywords}
@@
 A typical bicycle wheel (Fig.
-\ref{fig:schematic} consists of a slender rim connected to a hub by means of a system of very slender metal spokes.
+\ref{fig:schematic}) consists of a slender rim connected to a hub by means of a system of very slender metal spokes.
 The spokes are usually threaded at the rim side into spoke nipples which are inserted through the rim from the outside.
@@
 \includegraphics[scale=1.0]{figures/bmd_figures-06.eps}
-\caption{Deformation of a rim cross-section. The shear center $S\rightarrow S'$ translates by the vector $\bm{u} = u\e + v\ee + w\eee$ and rotates through an angle $\phi$.}
+\caption{Rim cross-section in the undeformed and deformed configuration. The shear center $S\rightarrow S'$ translates by the vector $\bm{u} = u\e + v\ee + w\eee$ and rotates through an angle $\phi$.}
 \label{fig:def}
@@
 
-The in-plane curvature $\kappa_1$ is assumed to remain constant during flexural-torsional buckling. The in-plane curvature, out-of-plane curvature and twist rate are
+The in-plane curvature $\kappa_1$ remains constant during flexural-torsional buckling. The in-plane curvature, out-of-plane curvature and twist rate are
 	\begin{equation}\label{eq:k1}
@@
 	\end{equation}
-The symbol $()'$ indicates a derivative with respect to $s$.
+The symbol $()'$ indicates a derivative with respect to arc-length $s$.
 It can easily be verified that Equations \ref{eq:k2} and \ref{eq:k3} simplify to the standard equations for straight beams in the limit $R\rightarrow \infty$.
-The curvature and twist rate produce a bending moment and twisting moment equal to
-	\begin{equation}\label{eq:M2}
-	M_2 = EI_{11} \kappa_2
+The total strain energy is decomposed into lateral bending, uniform torsion, and warping terms which are related to the curvatures.
+% The curvature and twist rate produce a bending moment and twisting moment equal to
+	% \begin{equation}\label{eq:M2}
+	% M_2 = EI_{11} \kappa_2
+	% \end{equation}
+	% \begin{equation}\label{eq:M3}
+	% M_3 = GJ \kappa_3 + EI_w \kappa_3''
+	% \end{equation}
+	\begin{equation}\label{eq:Urim}
+	U_{rim} = \frac{1}{2} \int_0^{2\pi} \left( EI_{11} \kappa_2^2 + GJ \kappa_3^2 + EI_w (\kappa_3')^2 \right)\, ds
 	\end{equation}
-	\begin{equation}\label{eq:M3}
-	M_3 = GJ \kappa_3 + EI_w \kappa_3''
-	\end{equation}
-The total strain energy is decomposed into lateral bending, uniform torsion, and warping terms which are related to the curvatures.
-	\begin{equation}\label{eq:Urim}
-	U_{rim} = \int_0^{2\pi} \left( \frac{1}{2} EI_{22} \kappa_2^2 + \frac{1}{2} GJ \kappa_3^2 + \frac{1}{2} EI_w (\kappa_3')^2 \right)\, ds
-	\end{equation}
-After buckling, a differential line segment $ds$ deforms to $dS$. The apparent shortening of the line segment along the $s$ direction is then
+The length of a line segment of the rim is assumed to remain constant during buckling.
+The apparent shortening of an initial line segment $ds$ is given by the difference between $ds$ and $ds'$, the in-plane component of the deformed line segment.
 	\begin{equation}\label{eq:ds}
-	ds - dS = ds - \sqrt{ds^2 - \left(\frac{du}{ds}ds\right)^2} \approx \frac{1}{2} (u')^2 \, ds
+	ds - ds' = ds - \sqrt{ds^2 - \left(\frac{du}{ds}ds\right)^2} \approx \frac{1}{2} (u')^2 \, ds
 	\end{equation}
@@
 In this paper we consider bicycle wheels with slender prestressed spokes.
-We model a single spoke as an elastic bar pinned at each end.
-This ensures that the spoke force is always collinear with the spoke axis.
+We model a single spoke as an elastic bar pinned at each end, ensuring that the spoke force is always collinear with the spoke axis.
 We fix our global coordinate system to the hub, which is assumed to be rigid.
-Therefore, the displacement of the hub node of a spoke is zero.
 The linearized elongation of a spoke is given by
@@
 	\end{equation}
-where \nn{} and \nnn{} are two orthogonal unit vectors orthogonal to \n{}, and $l$ is the spoke length.
-
-The elongation produces a net force on the rim parallel to the original spoke axis equal to
+where \nn{} and \nnn{} are two orthogonal unit vectors orthogonal to \n{}, and $l$ is the spoke length. The elongation produces a net force on the rim parallel to the original spoke axis equal to
 	\begin{equation}\label{eq:sF1}
@@
 The tension stiffness is the component responsible for the transverse vibrations of a guitar string, e.g.
+The term $\n\n$ is the dyadic product, or tensor product (also called the outer product), and is defined by the property $\n\n\cdot u = (\n\cdot u)\n$.
 
-Following the approach of Smith~\cite{Smith1901a} and Pippard~\cite{Pippard1932d}, we will approximate the spoke system as a continuum which exerts a distributed load on the rim.
+Formalizing the approach of Smith~\cite{Smith1901a} and Pippard~\cite{Pippard1932d}, we will approximate the spoke system as a continuum which exerts a distributed load on the rim.
 The stiffness per unit length is defined by the sum of the spoke stiffnesses in one periodic grouping of $n_p$ spokes, divided by the arc length of the grouping.
@@
 	\begin{equation}\label{eq:w_spokes}
-	\bar{w} = \int_0^{u_n} \bar{\bm{f}} \cdot d\bm{u}_n = \frac{1}{2} \bm{u}_n \cdot \bar{\bm{k}} \cdot \bm{u}_n
+	\bar{w} = \int_0^{\bm{u}_n} \bar{\bm{f}} \cdot d\bm{u}_n = \frac{1}{2} \bm{u}_n \cdot \bar{\bm{k}} \cdot \bm{u}_n
 	\end{equation}
@@
 	\end{equation}
-The wheel becomes unstable when the increase in strain energy is exactly balanced by the virtual work for a perturbing displacement $\delta u_n, \delta\phi_n$, i.e.
+The wheel becomes unstable when the increase in strain energy is exactly balanced by the virtual work for a perturbing displacement $\delta u, \delta\phi$, i.e.
 when the total potential energy is zero.
@@
 	\begin{multline}\label{eq:ECrit}
-	\frac{\pi EI_{11}}{R^3}(n^2 u_n - R\phi_n)^2 + \frac{\pi n^2}{R^2}\left(GJ + \frac{EI_w}{R^2}n^2\right)(u_n-R\phi_n)^2 + \\
-	\pi R(\bar{k}_{uu}^{el} + T\bar{k}_{uu}^{el})u_n^2 - N_r\frac{\pi n^2}{R}u_n^2=0
+	\frac{\pi EI_{11}}{R^3}(n^2 \delta u_n - R\delta\phi_n)^2 + \frac{\pi n^2}{R^2}\left(GJ + \frac{EI_w}{R^2}n^2\right)(\delta u_n-R\delta\phi_n)^2 + \\
+	\pi R(\bar{k}_{uu}^{el} + T\bar{k}_{uu}^{el})\delta u_n^2 - N_r\frac{\pi n^2}{R}\delta u_n^2=0
 	\end{multline}
-The torsional rigidity and warping rigidity can be combined into an effective torsional stiffness $\widetilde{GJ} = GJ + n^2EI_w/R^2$.
-Equation \ref{eq:ECrit} has a quadratic form and can be written using matrix notation:
+The torsional stiffness and warping stiffness can be combined into an effective torsional stiffness $\widetilde{GJ} = GJ + n^2EI_w/R^2$.
+Equation \ref{eq:ECrit} has a quadratic form and can be written in matrix notation:
 	\begin{equation}\label{eq:Qform}
 	\begin{bmatrix}
-	u_n\\\phi_n
+	\delta u_n\\\delta\phi_n
 	\end{bmatrix} \cdot
@@
 	\begin{bmatrix}
-	u_n\\\phi_n
+	\delta u_n\\\delta\phi_n
 	\end{bmatrix}
@@
 	\begin{equation}\label{eq:Ktorsion}
-	K_{torsion} = \widetilde{GJ}n^2\frac{\pi (n^2-1)^2}{R^3}
+	K_{torsion} = \frac{\pi\widetilde{GJ}n^2}{R^3}(n^2-1)^2
 	\end{equation}
 	\begin{equation}\label{eq:Kbending}
-	K_{torsion} = EI_{11}\frac{\pi (n^2-1)^2}{R^3}
+	K_{torsion} = \frac{\pi EI_{11}}{R^3}(n^2-1)^2
 	\end{equation}
@@
 
-With this definition, Eqn. \ref{eq:Tc_nondim} has an extremely simple form
+With this definition, the critical buckling tension for the $n$th mode (Eqn. \ref{eq:Tc_nondim}) has an extremely simple form
 	\begin{equation}\label{eq:Tc_Kn}
@@
 $\nn,\nnn$	& Two unit vectors perpendicular to the spoke axis\\
-$\e,\ee,\eee$ & Coordinate axes defined at the rim shear center\\
-$\bm{u}$	& Displacement of the rim shear center\\
-$u,v,w$		& Vector components of $\bm{u}$ in the \e{}, \ee{}, \eee{} directions.\\
-$\phi$		& Rotation of the rim cross-section about the $\hat{\bm{e}}_3$ axis.\\
-$\delta u, \delta \phi$ & Virtual displacement and rotation\\
-$u_n, \phi_n$ & Buckling mode coefficients\\
-$n$			& Buckling mode number\\
-$T$			& Tension in a spoke\\
-$T_{cn}$	& Critical tension for the $n$th mode\\
-$T_c$		& Critical buckling tension (minimum of $T_{cn}$ over all modes)\\
-$T_e$		& ``Euler'' tension--tension which would produce an axial force in the rim
-			  equal to the buckling load for a straight beam-column of length $2\pi R$\\
 $\bm{f}$	& Force exerted by a spoke on the rim\\
+$\bar{\bm{f}}$& Distributed force (per-unit-length) exerted by the spokes on the rim\\
 $f_1, f_2,f_3$ & Vector components of $\bm{f}$\\
@@
 $\bar{k}_{ij}$ & Tensor components of $\bar{\bm{k}}$\\
+$\e,\ee,\eee$ & Coordinate axes defined at the rim shear center\\
+$s$			& Arc-length coordinate along the rim\\
+$\bm{u}$	& Displacement of the rim shear center\\
+$\kappa_1,\kappa_2,\kappa_3$ & Spatial curvature of the rim\\
+$u,v,w$		& Vector components of $\bm{u}$ in the \e{}, \ee{}, \eee{} directions.\\
+$\phi$		& Rotation of the rim cross-section about the $\hat{\bm{e}}_3$ axis.\\
+$\delta u, \delta \phi$ & Virtual displacement and rotation\\
+$u_n, \phi_n$ & Buckling mode coefficients\\
+$n$			& Buckling mode number\\
+$U_{rim}$	& Strain energy in the rim due to buckling\\
+$V_{rim}$	& Virtual work of internal forces in the rim\\
+$U_{spokes}$& Strain energy in the spokes due to buckling\\
+$T$			& Tension in a spoke\\
+$T_{cn}$	& Critical tension for the $n$th mode\\
+$T_c$		& Critical buckling tension (minimum of $T_{cn}$ over all modes)\\
+$T_e$		& ``Euler'' tension--tension which would produce an axial force in the rim
+			  equal to the buckling load for a straight beam-column of length $2\pi R$\\
 $R$			& Rim radius--measured from the rim shear center\\
@@
 $\tilde{\mu}$ & Ratio of effective torsional rigidity to bending rigidity: $\widetilde{GJ}/EI_{11}$\\
-$\lambda$	& Ratio of spoke system stiffness to rim bending stiffness OR ratio of rim radius to 
-			  the characteristic length of a beam on an elastic foundation.\\
+$\lambda$	& Ratio of spoke system lateral stiffness to rim bending stiffness OR ratio
+			  of rim radius to the characteristic length of a beam on an elastic foundation.\\
 \hline

### 2016-08-20 latex/paper.tex
Conclusion and minor edits

@@
 \section{INTRODUCTION}
+The earliest ``bicycles'' were built in the workshops of carriage makers who adapted the wheel technology of the time.
+These wheels typically had iron tires, wooden rims and stout wooden spokes which bore the weight of the machine and rider in compression.
+The spokes needed to be wide enough to prevent lateral buckling, which made them very heavy and limited their diameter.
+Around 1869, the tension-spoke wheel, an idea proposed in 1808 by Sir George Cayley but never realized, found its first application on the rapidly-developing bicycle \cite{Ackroyd2011a,Clayton1991a}.
+Tangential-spoking was developed soon thereafter, and the typical bicycle wheel has not changed dramatically in form since then.
 
-The bicycle wheel is a prestressed structure and is susceptible to buckling under internal forces.
+The tension-spoke wheel is a prestressed structure and is susceptible to buckling under internal forces.
 As the spokes are tightened uniformly, the rim deforms radially to accommodate the spoke strain.
-At a critical tension, the system reaches a bifurcation point and the rim buckles out of its initial plane.
-The post-buckling configuration is generally stable and the original shape of the wheel can be recovered by reducing the tension.
+At a critical tension, the rim buckles out of its initial plane.
+The post-buckling configuration is generally stable and the original shape of the wheel can often be recovered by reducing the tension.
 Buckling can also be triggered by external forces in an otherwise stable wheel leading to a release of strain energy.
@@
 He did not discuss the problem further, but suggested that wheel failure commonly occurs due to a loss of lateral stability caused by spoke buckling.
-Pippard and Francis~\cite{Pippard1932d} derived a model for lateral stiffness based on the elastic foundation model, but did not discuss stability and neglected any effects of spoke tension.
+Pippard and Francis~\cite{Pippard1932d} derived an equation for lateral deflection based on an elastic foundation model, but did not discuss stability and neglected any effects of spoke tension.
 
@@
 The problem of the prestressed bicycle wheel is unique for a number of reasons.
-First, the buckling load is internal to the structure.
+First, buckling is caused by internal rather than external forces.
 Second, the spokes act both as elastic restraints resisting buckling and as prestressing elements causing buckling.
@@
 Large observation wheels such as the London Eye~\cite{Mann2001a} and the Singapore Flyer~\cite{Allsop2009a} resemble large bicycle wheels and achieve lateral stability due to the bracing angle of prestressed cables, and must be designed against flexural-torsional buckling.
-
-Here we derive a simple solution for buckling of a spoked bicycle wheel under self-tension and investigate several special cases.
 
@@
 
-To hold the spokes under tension, the rim must be under compression.
+To hold the spokes in tension, the rim must be under compression.
 An accurate approximation\footnote{95\% accurate for 6 spokes, $\alpha=10^{\circ}$, $\beta=10^{\circ}$.
@@
 For simplicity we will assume that the shear center and centroid of the cross-section coincide.
-Except for warping deformation, plane cross-sections remain planar and shear flexibility of the cross-section is neglected.
-Furthermore, we assume that the rim radius is much larger than the height of the rim profile.
+Except for warping deformation, plane cross-sections remain planar and transverse shear flexibility is neglected.
+Furthermore, we assume that the rim radius is much larger than the cross-sectional dimensions.
 
@@
 
-The in-plane curvature $\kappa_1$ remains constant during flexural-torsional buckling. The in-plane curvature, out-of-plane curvature and twist rate are
+The deformed configuration of the rim is described by the displacement of the shear center $\bm{u}=u\e + v\ee + w\eee$ and the twist angle $\phi$.
+The in-plane curvature $\kappa_1$ remains constant during flexural-torsional buckling.
+The in-plane curvature, out-of-plane curvature and twist rate are
 	\begin{equation}\label{eq:k1}
@@
 The symbol $()'$ indicates a derivative with respect to arc-length $s$.
-It can easily be verified that Equations \ref{eq:k2} and \ref{eq:k3} simplify to the standard equations for straight beams in the limit $R\rightarrow \infty$.
+It can easily be verified that Equations \ref{eq:k1}--\ref{eq:k3} simplify to the standard equations for straight beams in the limit $R\rightarrow \infty$.
 The total strain energy is decomposed into lateral bending, uniform torsion, and warping terms which are related to the curvatures.
@@
 	\end{equation}
-The axial compressive force in the rim moves through this differential displacement, performing virtual work equal to
+The axial compressive force in the rim moves through this displacement, performing virtual work equal to
 	\begin{equation}\label{eq:Vrim}
@@
 We model a single spoke as an elastic bar pinned at each end, ensuring that the spoke force is always collinear with the spoke axis.
-We fix our global coordinate system to the hub, which is assumed to be rigid.
+The hub is assumed to be rigid.
 The linearized elongation of a spoke is given by
@@
 	\end{equation}
-where $\bm{u}_n$ is the displacement vector of the spoke nipple and \n{} is a unit vector pointing from the spoke nipple to the hub eyelet. The linearized rotation of the spoke is given by
+where $\bm{u}_n$ is the displacement vector of the spoke nipple and \n{} is a unit vector pointing along the axis of the spoke. The linearized rotation of the spoke is given by
 	\begin{equation}\label{eq:srot1}
@@
 	\begin{equation}\label{eq:sF3}
-	f_3 = T \sin{\Omega_2} \approx \frac{T}{l} \nn \cdot \bm{u}_n
+	f_3 = T \sin{\Omega_2} \approx \frac{T}{l} \nnn \cdot \bm{u}_n
 	\end{equation}
@@
 The tension stiffness is the component responsible for the transverse vibrations of a guitar string, e.g.
-The term $\n\n$ is the dyadic product, or tensor product (also called the outer product), and is defined by the property $\n\n\cdot u = (\n\cdot u)\n$.
+The term $\n\n$ is the dyadic product, or tensor product (also called the outer product), and is defined by the property $\n\n\cdot \bm{v} = (\n\cdot \bm{v})\n$ for any vector $\bm{v}$.
 
-Formalizing the approach of Smith~\cite{Smith1901a} and Pippard~\cite{Pippard1932d}, we will approximate the spoke system as a continuum which exerts a distributed load on the rim.
-The stiffness per unit length is defined by the sum of the spoke stiffnesses in one periodic grouping of $n_p$ spokes, divided by the arc length of the grouping.
+Formalizing the approach of Smith~\cite{Smith1901a} and Pippard~\cite{Pippard1932d}, we will approximate the spoke system as a continuous elastic ``disc'' which exerts a distributed load on the rim.
+The stiffness per unit length is given by the sum of the spoke stiffnesses in one periodic grouping of $n_p$ spokes, divided by the arc length of the grouping.
 	\begin{equation}\label{eq:kbar}
@@
 	\end{equation}
-The continuum spoke stiffness can be written $\bar{\bm{k}} = \bar{\bm{k}}^{el} + T\bar{\bm{k}}^{tens}$, where the dependence on $T$ is shown explicitly.
+The continuum spoke stiffness can be written $\bar{\bm{k}} = \bar{\bm{k}}^{el} + \bar{\bm{k}}^{tens}$, where the second term is proportional to $T$.
 A very close approximation for the $\bar{k}_{uu}$ component for a symmetrically dished wheel is given by
@@
 \subsection{Total potential energy and buckling criterion}
-
-The buckled shape of the wheel must be periodic, so we assume a buckled shape of the form
+The wheel becomes unstable when the increase in strain energy is exactly balanced by the virtual work for a perturbing displacement $\delta u, \delta\phi$, i.e.
+when the total potential energy is zero.
+	\begin{equation}\label{eq:TotPot}
+	\Pi = U_{rim} + U_{spokes} - V_{rim} = 0
+	\end{equation}
+We assume a buckled shape of the form
 	\begin{equation}\label{eq:modeshape}
@@
 	\end{split}
-	\end{equation}
-The wheel becomes unstable when the increase in strain energy is exactly balanced by the virtual work for a perturbing displacement $\delta u, \delta\phi$, i.e.
-when the total potential energy is zero.
-	\begin{equation}\label{eq:TotPot}
-	\Pi = U_{rim} + U_{spokes} - V_{rim} = 0
 	\end{equation}
@@
 	\frac{\pi EI_{11}}{R^3}(n^2 \delta u_n - R\delta\phi_n)^2 + \frac{\pi n^2}{R^2}\left(GJ + \frac{EI_w}{R^2}n^2\right)(\delta u_n-R\delta\phi_n)^2 + \\
-	\pi R(\bar{k}_{uu}^{el} + T\bar{k}_{uu}^{el})\delta u_n^2 - N_r\frac{\pi n^2}{R}\delta u_n^2=0
+	\pi R(\bar{k}_{uu}^{el} + \bar{k}_{uu}^{el})\delta u_n^2 - N_r\frac{\pi n^2}{R}\delta u_n^2=0
 	\end{multline}
@@
 Expanding the determinant in Eqn.
-\ref{eq:Qform} and substituting Eqn.
-\ref{eq:TN} yields a linear equation for the critical buckling tension of the $n$th mode.
+\ref{eq:Qform} and substituting Eqns.
+\ref{eq:TN} and \ref{eq:kuu} yields a linear equation for the critical buckling tension of the $n$th mode.
 	\begin{multline}\label{eq:Tc_eq}
-	\left(\frac{2\pi EI_{22}}{R}+\frac{2\pi \widetilde{GJ}n^2}{R} \right) \left( \frac{2\pi \widetilde{GJ}n^2}{R^3} + \frac{2\pi EI_{22}n^4}{R^3} + 2\pi R\bar{k}_{uu}^{el} + \frac{n_sT_{cn}}{l} - \frac{n_sn^2T_{cn}}{R}\right)\\
-	- \left( \frac{2\pi EI_{22}n^2}{R^2} + \frac{2\pi \widetilde{GJ}n^2}{R^2} \right)^2 = 0
+	\left(\frac{2\pi EI_{11}}{R}+\frac{2\pi \widetilde{GJ}n^2}{R} \right) \left( \frac{2\pi \widetilde{GJ}n^2}{R^3} + \frac{2\pi EI_{11}n^4}{R^3} + 2\pi R\bar{k}_{uu}^{el} + \frac{n_sT_{cn}}{l} - \frac{n_sn^2T_{cn}}{R}\right)\\
+	- \left( \frac{2\pi EI_{11}n^2}{R^2} + \frac{2\pi \widetilde{GJ}n^2}{R^2} \right)^2 = 0
 	\end{multline}
@@
 	\begin{equation}\label{eq:Tc_nondim}
-	\frac{T_{cn}}{T_e} = \left( \frac{\tilde{u}n^2(n^2-1)^2}{1+\tilde{\mu}n^2} + \lambda \right) \left(\frac{1}{n^2-1} \right)
+	\frac{T_{cn}}{T_e} = \left( \frac{\tilde{u}n^2(n^2-1)^2}{1+\tilde{\mu}n^2} + \lambda \right) \left(\frac{1}{n^2-R/l} \right)
 	\end{equation}
-where the ``Euler tension'' $T_e=2\pi EI_{22}/n_sR^2$ is the tension that would produce a compressive force in the rim equal to the classical buckling load for a straight beam-column of length $2\pi R$.
-The parameter $\tilde{\mu} = \widetilde{GJ}/EI_{22}$ is the ratio of the effective torsional stiffness to the lateral bending stiffness.
-The parameter $\lambda = \bar{k}_{uu}^{el}R^4/EI_{22}$ is the ratio of the spoke system stiffness $\bar{k}_{uu}^{el}R$ to the rim bending stiffness $EI_{22}/R^3$.
+\,\,\,\, where the ``Euler tension'' $T_e=2\pi EI_{11}/n_sR^2$ is the tension that would produce a compressive force in the rim equal to the classical buckling load for a straight beam-column of length $2\pi R$.
+The parameter $\tilde{\mu} = \widetilde{GJ}/EI_{11}$ is the ratio of the effective torsional stiffness to the lateral bending stiffness.
+The parameter $\lambda = \bar{k}_{uu}^{el}R^4/EI_{11}$ is the ratio of the spoke system stiffness $\bar{k}_{uu}^{el}R$ to the rim bending stiffness $EI_{11}/R^3$.
+For most wheels, $R/l\approx 1$.
 
@@
 \includegraphics[scale=1.0]{figures/bmd_figures-04.eps}
-\caption{\textbf{(a)} Buckling mode parameter map.
-Only the first 4 modes are shown.
+\caption{\textbf{(a)} Buckling mode parameter map showing the lowest-energy buckling mode in each region.
 \textbf{(b)}--\textbf{(c)} Normalized buckling tension vs.
 $\tilde{\mu}$ and $\lambda$.
-Solid black line = Eqn.
-\ref{eq:Tc_nondim}, dashed red line = power law, blue stars = finite-element simulations.
-In (a) $\lambda=10$ while in (b) $\tilde{\mu} = 0.38$.}
+Solid black line: Eqn.
+\ref{eq:Tc_nondim}, dashed red line: power law, blue stars = finite-element calculations.
+(a) $\lambda=10$, (b) $\tilde{\mu} = 0.38$.}
 \label{fig:Tc_nondim}
@@
 
-Since $n$ is a discrete variable, there is no closed-form for the minimum $T_{cn}$ for a given $(\tilde{\mu},\lambda)$.
+Since $n$ is a discrete variable, there is no closed-form expression for the minimum $T_{cn}$ for a given $(\tilde{\mu},\lambda)$.
+The minimum-energy buckling mode must be found by comparing $T_{cn}$ for the first several modes and taking the smallest value (see Fig.
+\ref{fig:Tc_nondim} (a)).
 However, two important \textit{approximate} solutions can be derived by replacing the discrete $n$ with a continuous variable $\bar{n}$ and minimizing Eqn.
@@
 \subsection{Rims with low torsional stiffness}\label{sec:powerlaw_1}
-Many bicycle rims are built from thin-walled open channels which have very low torsional stiffness compared with their bending stiffness.
+Many bicycle rims are built from thin-walled open channel sections which have very low torsional stiffness compared with their bending stiffness.
 Replacing the discrete mode number $n$ with a continuous variable $\bar{n}$, we adopt the following ansatz: (a) $\bar{n}^2 \gg 1$ and (b) $\tilde{\mu}\bar{n}^2 \ll 1$.
@@
 	\begin{equation}\label{eq:nbar_small_mu}
-	\bar{n} = \left(\frac{\lambda}{2\tilde{\mu}} \right)
+	\bar{n} = \left(\frac{\lambda}{2\tilde{\mu}} \right)^{1/6}
 	\end{equation}
-Substituting Eqn. \ref{eq:nbar_small_mu} into Eqn. \ref{eq:Tc_small_mu} yields, after simplification, a simple power law relationship.
+Substituting Eqn. \ref{eq:nbar_small_mu} into Eqn. \ref{eq:Tc_small_mu} yields, after simplification, a power law for $T_c$.
 	\begin{equation}\label{eq:Tc_small_mu}
@@
 Equation \ref{eq:Tc_small_mu} is shown in Fig.
-\ref{fig:Tc_nondim} next to Equation \ref{eq:Tc_nondim} and finite-element calculations.
-The power law is strictly less than Equation \ref{eq:Tc_nondim} which conveniently gives a conservative design criterion.
+\ref{fig:Tc_nondim} next to Eqn.
+\ref{eq:Tc_nondim} and finite-element calculations.
+The power law is strictly less than Eqn.
+\ref{eq:Tc_nondim} which conveniently gives a conservative design criterion.
 
@@
 Substituting Eqn.
-\ref{eq:TN} for $T_c$ gives $N_{rc}=2\sqrt{\bar{k}_{uu}^{el}EI_{11}}$, which is precisely the buckling load for an infinite, straight beam on an elastic foundation~\cite{Hetenyi1946a}.
+\ref{eq:TN} for $T_c$ gives $N_{rc}=2\left(\bar{k}_{uu}^{el}EI_{11}\right)^{1/2}$, which is precisely the buckling load for an infinite, straight beam on an elastic foundation~\cite{Hetenyi1946b}.
 Figure \ref{fig:Tc_nondim} (c) shows a comparison of Eqn.
@@
 \ref{eq:modeshape}.
-Since the Fourier modes are orthogonal, the total strain energy is the sum of the strain energy in each mode.
+Since the Fourier modes are orthogonal, the total strain energy is the sum of the strain energies in each mode.
 	\begin{equation}\label{eq:Umode}
@@
 	\begin{equation}\label{eq:Kspokes}
-	K_{spokes} = \pi R \bar{k}_{uu}
+	K_{spokes} = \pi R \bar{k}_{uu}^{el}
 	\end{equation}
@@
 The rim was interpolated with sufficient elements such that any given beam segment spanned less than 6$^{\circ}$ of the circle.
+Each spoke was modeled with 10 beam elements to accurately reproduce local spoke buckling.
 
@@
 \section{CONCLUSIONS}
-...
+The bicycle wheel relies on spoke tension to support large loads in the radial direction.
+As a rule-of-thumb, a typical wheel can withstand a load of approximately $2T$ before the bottom-most spokes lose tension.
+Therefore, higher tension is desirable to increase the radial strength of the wheel.
+However, we have shown that if the spoke tension is too high, the rim becomes unstable under internal forces, without even considering external loads.
+Tension---the property which gives a wheel its strength---can also be its downfall.
+
+Our theory points towards some practical design rules: \textbf{(1)} Since the rim bending stiffness and torsional stiffness effectively act like springs in series, the more flexible mode will dominate.
+Therefore they should be in relative balance.
+Modern double-walled rims are advantageous in this respect because of their closed cross-section.
+\textbf{(2)} Since the rim and spoke stiffnesses act like springs in parallel, they can both be increased to roughly equal effect.
+\textbf{(3)} Since the lateral spoke stiffness scales roughly as $\sin^2{\alpha}$, the most significant factor affecting the spoke stiffness is the cone angle $\alpha$.
+Even a modest increase in hub width can increase the stiffness and buckling tension dramatically, while having negligible effect on the radial stiffness.
+
+One might wonder, why do practically all wheels fail in the $n=2$ mode, when higher modes are possible?
+We offer three speculative possibilities: \textbf{(1)} the geometry and stiffness of wheels are such that most wheels fall into the $n=2$ regime of Fig.
+\ref{fig:Tc_nondim}.
+\textbf{(2)} Wheels which would buckle into higher modes typically fail in other manners first (such as spoke yielding or nipple fracture).
+\textbf{(3)} Rim failure is usually accompanied by spoke buckling, which reduces the effective wheel stiffness and in turn favors the $n=2$ mode.
+Preliminary non-linear finite-element simulations of wheel failure under external loads provide some evidence for (3), but more work is required to apply our analysis to real-world wheels.
 
@@
 \section*{LIST OF TERMS}
-The following notation is used in this paper:
+A partial list of symbols is given below. Other symbols are defined in-place on not re-used.
 
@@
 $\bar{\bm{k}}$ & Equivalent stiffness per-unit-length of the spoke system. $\bar{\bm{k}} = \bar{\bm{k}}^{el} + \bar{\bm{k}}^{tens}$\\
-$\bar{\bm{k}}^{el}$ & Part of $\bar{\bm{k}}$ proportional to $E_sA_s$\\
-$\bar{\bm{k}}^{tens}$ & Part of $\bar{\bm{k}}$ proportional to $T$\\
+$\bar{\bm{k}}^{el},\bar{\bm{k}}^{tens}$ & Part of $\bar{\bm{k}}$ proportional to $E_sA_s$ and $T$, respectively\\
 $\bar{k}_{ij}$ & Tensor components of $\bar{\bm{k}}$\\
@@
 $s$			& Arc-length coordinate along the rim\\
+$\kappa_1,\kappa_2,\kappa_3$ & Spatial curvature of the rim\\
 $\bm{u}$	& Displacement of the rim shear center\\
-$\kappa_1,\kappa_2,\kappa_3$ & Spatial curvature of the rim\\
 $u,v,w$		& Vector components of $\bm{u}$ in the \e{}, \ee{}, \eee{} directions.\\
@@
 $\delta u, \delta \phi$ & Virtual displacement and rotation\\
-$u_n, \phi_n$ & Buckling mode coefficients\\
+$\delta u_n, \delta\phi_n$ & Buckling mode coefficients\\
 $n$			& Buckling mode number\\
@@
 $T$			& Tension in a spoke\\
-$T_{cn}$	& Critical tension for the $n$th mode\\
+$T_{cn}$	& Critical spoke tension for the $n$th mode\\
 $T_c$		& Critical buckling tension (minimum of $T_{cn}$ over all modes)\\

### 2016-08-31 latex/paper.tex
Removed dynamic collapse from title
First round of edits from Jim

@@
 \begin{center}
-\fontsize{14}{20}{\bf Buckling and dynamic collapse of the bicycle wheel}
+\fontsize{14}{20}{\bf Buckling of the bicycle wheel}
 \end{center}
@@
 360 Huntington Ave., Boston, MA 02115, USA\\
-e-mail: j.papadopoulos@neu.edu\\[2.5ex]
+e-mail: j.papadopoulos@northeastern.edu\\[2.5ex]
 
@@
 \section*{ABSTRACT}
-The spokes of a bicycle wheel must be kept under tension to prevent them from buckling under external loads.
-However, above a certain critical tension, the rim will bend and twist out of its plane into a ``pringle'' or ``taco'' shape.
+Bicycle spokes brace and stiffen the slender rim, but only as long as they do not lose pre-tension under external loads.
+However, above a certain critical tension, the rim will buckle into a non-planar saddle shape.
 Despite its importance for wheel strength and safety, there is no formula in the literature for the maximum tension a wheel can withstand before buckling.
@@
 Here we derive such a formula.
-The non-dimensionalized buckling tension depends only on two parameters which are identified as the ratio of torsional stiffness to bending stiffness, and the ratio of spoke stiffness to bending stiffness.
-The number of ``waves'' in the buckled shape is determined by these two stiffness ratios and may in general by greater than 2.
-Finally, we show that the wheel may be represented by a system of equivalent springs: two series-connected springs for the rim bending and torsion stiffness, connected in parallel with a spring for the spoke system.
+The non-dimensionalized buckling tension depends only on two parameters which are identified as the ratio of torsional stiffness to bending stiffness, and the ratio of lateral spoke bracing stiffness to rim bending stiffness.
+The precise buckled shape is determined by these two stiffness ratios and the number of positive peaks of the lateral displacement may in general by greater than 2.
+We show that the wheel may be represented by a system of equivalent springs: two series-connected springs for the rim bending and torsion stiffness, connected in parallel with a spring for the spoke system.
 
@@
 bicycle wheel, 
+bicycle safety, 
+wheel strength, 
 buckling, 
-bifurcation,
-bicycle safety.
+flexural-torsional.
 \end{keywords}
@@
 \section{INTRODUCTION}
-The earliest ``bicycles'' were built in the workshops of carriage makers who adapted the wheel technology of the time.
-These wheels typically had iron tires, wooden rims and stout wooden spokes which bore the weight of the machine and rider in compression.
-The spokes needed to be wide enough to prevent lateral buckling, which made them very heavy and limited their diameter.
-Around 1869, the tension-spoke wheel, an idea proposed in 1808 by Sir George Cayley but never realized, found its first application on the rapidly-developing bicycle \cite{Ackroyd2011a,Clayton1991a}.
+The earliest bicycles were built in the workshops of carriage makers who adapted the wheel technology of the time.
+These wheels typically had iron tires, wooden rims and stout wooden spokes which bore the weight of the machine and rider in compression~\cite{Herlihy2004a}.
+The spokes needed to be wide enough to prevent lateral buckling and to keep the rim planar, which made them very heavy and limited their diameter.
+Around 1869, the tension-spoke wheel, an idea proposed in 1808 by Sir George Cayley but never realized, found its first application on the rapidly-developing bicycle~\cite{Ackroyd2011a,Clayton1991b}.
 Tangential-spoking was developed soon thereafter, and the typical bicycle wheel has not changed dramatically in form since then.
@@
 The tension-spoke wheel is a prestressed structure and is susceptible to buckling under internal forces.
-As the spokes are tightened uniformly, the rim deforms radially to accommodate the spoke strain.
+As the spokes are tightened uniformly, the rim deforms radially under uniform compression.
 At a critical tension, the rim buckles out of its initial plane.
-The post-buckling configuration is generally stable and the original shape of the wheel can often be recovered by reducing the tension.
+The post-buckling load-displacement curve is generally stable and the original shape of the wheel can often be recovered by reducing the tension.
 Buckling can also be triggered by external forces in an otherwise stable wheel leading to a release of strain energy.
-This is an unstable process which generally leads to a permanently ``taco-ed'' wheel.
+This process can lead to plastic yielding of the rim which makes wheel repair significantly more difficult.
 
@@
 
-He did not discuss the problem further, but suggested that wheel failure commonly occurs due to a loss of lateral stability caused by spoke buckling.
+He did not discuss the problem of buckling under tension further, but proposed that wheel failure commonly occurs due to lateral buckling when external loads have caused spokes to lose tension.
 Pippard and Francis~\cite{Pippard1932d} derived an equation for lateral deflection based on an elastic foundation model, but did not discuss stability and neglected any effects of spoke tension.
 
-The bicycle wheel can be modeled as an elastically supported arch in which the included angle is allowed to go to $2\pi$.
+The theory of bicycle wheel buckling is closely related to the theory of flexural-torsional buckling of arches.
 Timoshenko and Gere~\cite{Timoshenko1961a} gave a formula for the critical load for a ring with doubly-symmetric cross-section subjected to radial loads.
-The theory of flexural-torsional buckling of monosymmetric arches (bicycle rims have only one plane of symmetry) was broadly formalized by Trahair and Papangelis~\cite{Trahair1987b} using the virtual work approach to derive the equilibrium and stability equations.
+The theory of flexural-torsional buckling of monosymmetric arches (bicycle rim cross-sections generally have only one plane of symmetry) was broadly formalized by Trahair and Papangelis~\cite{Trahair1987b} using the virtual work approach to derive the equilibrium and stability equations.
 Their theory has been extended to treat arches restrained by continuous~\cite{Pi2002b} or discrete~\cite{Bradford2002d} elastic supports and elastic end restraints~\cite{Guo2014b}.
@@
 Third, the lateral, radial, and torsional restraining actions of the spokes are coupled---i.e.
-lateral deflection at a spoke may produce lateral, radial, and torsional reactions.
+lateral deflection of the rim may produce lateral, radial, and torsional reactions from the spokes.
 These considerations extend to other structural systems.
-Large observation wheels such as the London Eye~\cite{Mann2001a} and the Singapore Flyer~\cite{Allsop2009a} resemble large bicycle wheels and achieve lateral stability due to the bracing angle of prestressed cables, and must be designed against flexural-torsional buckling.
+Large observation wheels such as the London Eye~\cite{Mann2001a} and the Singapore Flyer~\cite{Allsop2009a} resemble large bicycle wheels with lateral stability due to the bracing angle of prestressed cables, and must be designed against flexural-torsional buckling.
 
@@
 To hold the spokes in tension, the rim must be under compression.
-An accurate approximation\footnote{95\% accurate for 6 spokes, $\alpha=10^{\circ}$, $\beta=10^{\circ}$.
+An accurate approximation\footnote{95\% accurate for 6 spokes, $\alpha=10^{\circ}$, $\beta=10^{\circ}$, where angles $\alpha$ and $\beta$ are defined in Fig.
+\ref{fig:schematic}.
 Increasing the number of spokes only improves the estimate.} was given by Sharp~\cite{Sharp1977a} by considering a force balance on the top (or bottom) half of the rim.
@@
 The rim is modeled as an initially circular beam with a constant cross-section.
-For simplicity we will assume that the shear center and centroid of the cross-section coincide.
+For simplicity in this abbreviated treatment, we will assume that the shear center and centroid of the cross-section coincide.
 Except for warping deformation, plane cross-sections remain planar and transverse shear flexibility is neglected.
@@
 	\end{equation}
-The symbol $()'$ indicates a derivative with respect to arc-length $s$.
+The symbol $()'$ indicates a derivative with respect to arc-length $s$ along the \eee{} direction.
 It can easily be verified that Equations \ref{eq:k1}--\ref{eq:k3} simplify to the standard equations for straight beams in the limit $R\rightarrow \infty$.
@@
 	\end{equation}
+The warping energy is primarily related to the bending energy in the sidewalls (which act as beam flanges) of the rim\footnote{See http://www.steel-insdag.org/TeachingMaterial/Chapter17.pdf for a simple treatment.}.
+The warping constant $I_w$ has units of m$^6$, and for single-wall rims the warping stiffness $EI_w$ generally exceeds the torsion stiffness $GJ$.
+
 The length of a line segment of the rim is assumed to remain constant during buckling.
@@
 Formalizing the approach of Smith~\cite{Smith1901a} and Pippard~\cite{Pippard1932d}, we will approximate the spoke system as a continuous elastic ``disc'' which exerts a distributed load on the rim.
-The stiffness per unit length is given by the sum of the spoke stiffnesses in one periodic grouping of $n_p$ spokes, divided by the arc length of the grouping.
+The stiffness per unit length is given by the sum of the spoke stiffnesses in one periodic grouping of $n_p$ spokes, divided by the arc length of the grouping\footnote{For a conventional cross-laced wheel, $n_p=4$: drive-side pulling spoke, drive-side trailing spoke, left-side pulling spoke, left-side trailing spoke.
+For a radial-spoked wheel, $n_p=2$: drive-side spoke, left-side spoke.}.
 	\begin{equation}\label{eq:kbar}
@@
 	\end{equation}
-The line of action is assumed to pass sufficiently close to the rim shear center that any difference between the displacement of the spoke nipple and the displacement of the shear center can be ignored.
+Each spoke is assumed to terminate at the rim shear center.
 Therefore $\bm{u}_n = \bm{u}$.
@@
 	\end{equation}
+\,\,\,\, where $n$ is the number of positive peaks of the lateral displacement.
 Substituting the buckling mode shape Eqn.
@@
 \subsection{Rims with low torsional stiffness}\label{sec:powerlaw_1}
-Many bicycle rims are built from thin-walled open channel sections which have very low torsional stiffness compared with their bending stiffness.
+Many bicycle rims are built from thin-walled open channel sections which have very low torsional stiffness $\widetilde{GJ}$ compared with their bending stiffness.
 Replacing the discrete mode number $n$ with a continuous variable $\bar{n}$, we adopt the following ansatz: (a) $\bar{n}^2 \gg 1$ and (b) $\tilde{\mu}\bar{n}^2 \ll 1$.
@@
 	\end{equation}
-Substituting Eqn. \ref{eq:nbar_small_mu} into Eqn. \ref{eq:Tc_small_mu} yields, after simplification, a power law for $T_c$.
+Substituting Eqn. \ref{eq:nbar_small_mu} into Eqn. \ref{eq:Tcn_small_mu} yields a power law for $T_c$.
 	\begin{equation}\label{eq:Tc_small_mu}
@@
 
-\subsection{Spokes much stiffer than the rim}
+\subsection{Lateral stiffness of spoke system much greater than rim bending stiffness}
 A second power law regime appears when $\lambda \gg 1$ and $\tilde{\mu} \sim 1$. Under these conditions, Eqn. \ref{eq:Tc_nondim} becomes
@@
 \section{EQUIVALENT SPRING MODEL}
+The form of Eqn.
+\ref{eq:Tc_nondim} suggests that the buckling tension is governed by an equivalent stiffness which combines the bending stiffness and torsion stiffness of the rim and the lateral stiffness of the spoke system.
+Here we will describe an equivalent system of linear springs (Fig.
+\ref{fig:Kn}) consisting of two springs connected in series representing the torsion and bending stiffnesses of the rim, in parallel with a single spring representing the lateral stiffness of the spoke system.
+
 The lateral stiffness of the wheel can be decomposed into orthogonal modes in the form of Eqn.
@@
 	\end{equation}
-% The stiffness of the $n$th mode is then defined as
-	% \begin{equation}\label{eq:Kn}
-	% K_n = P/u_n = \left(\frac{\pi}{R^3}\right) \frac{EI_{11}\widetilde{GJ}n^2(n^2-1)^2}{EI_{11}+\widetilde{GJ}n^2} + \pi R \bar{k}_{uu}
-	% \end{equation}
-
 Defining an equivalent torsion stiffness, bending stiffness, and spoke stiffness, the mode stiffness $P/u_n$ can be obtained from Eqn. \ref{eq:un} as
@@
 
-% \section{SUB-CRITICAL BUCKLING UNDER LATERAL LOADS}
-% ...
-
 \section{FINITE-ELEMENT CALCULATIONS}
@@
 The spokes and rim were both modeled using 2-node linear beam elements including shear flexibility (element type B31 in the ABAQUS library).
-The rim was interpolated with sufficient elements such that any given beam segment spanned less than 6$^{\circ}$ of the circle.
+The rim was interpolated with sufficient elements such that any given beam segment spanned less than 6$^{\circ}$ of the circle (e.g.
+2 rim elements between each spoke for a 36-spoke wheel).
 Each spoke was modeled with 10 beam elements to accurately reproduce local spoke buckling.
@@
 A controlled tensioning strain was applied to the spokes by defining a coefficient of thermal expansion of 1.0 for the spokes (but not the rim) and ramping the temperature down to 150\% of the expected critical buckling strain.
-Mesh imperfections of the form $\cos{ns/R}$ were added to perturb the initial lateral positions of the rim nodes.
-Modes starting with $n=2$ up to the expected buckling mode were included.
+The initial rim shape was given a geometric imperfection by perturbing the lateral coordinate of the rim nodes by $\cos{ns/R}$.
+Imperfections starting with $n=2$ up to the expected buckling mode were included.
 The solution was obtained via an implicit integration procedure (ABAQUS Standard) including non-linear geometric effects.
@@
 The bicycle wheel relies on spoke tension to support large loads in the radial direction.
-As a rule-of-thumb, a typical wheel can withstand a load of approximately $2T$ before the bottom-most spokes lose tension.
-Therefore, higher tension is desirable to increase the radial strength of the wheel.
+As a rule-of-thumb, a typical wheel can withstand a radial load of approximately $2T$ before the bottom-most spokes lose tension.
+Therefore, higher tension is desirable to increase the strength of the wheel.
 However, we have shown that if the spoke tension is too high, the rim becomes unstable under internal forces, without even considering external loads.
@@
 Therefore they should be in relative balance.
-Modern double-walled rims are advantageous in this respect because of their closed cross-section.
+Modern double-walled rims are advantageous in this respect because of their closed cross-section, while classic single-walled rims are much more flexible in torsion.
 \textbf{(2)} Since the rim and spoke stiffnesses act like springs in parallel, they can both be increased to roughly equal effect.
@@
 Even a modest increase in hub width can increase the stiffness and buckling tension dramatically, while having negligible effect on the radial stiffness.
+
+The spokes give a significant contribution to the buckling resistance of bicycle wheels.
+The spoke system lateral stiffness (Eqn.
+\ref{eq:kuu}) scales directly with the elastic modulus (material-dependent, but not alloy-dependent) and the cross-sectional area, and inversely with the spoke length (which varies only slightly depending on spoke pattern).
+It is counter-intuitive that increasing the number of spokes actually slightly \textit{decreases} the buckling tension (due to the factor of $n_s$ appearing in $T_e$ in Eqn.
+\ref{eq:Tc_nondim}).
+However, increasing the number of spokes also reduces the fraction of load transferred to a single spoke and thus reduces the required tension.
+The net effect is that more spokes results in a stronger wheel.
 
@@
 Preliminary non-linear finite-element simulations of wheel failure under external loads provide some evidence for (3), but more work is required to apply our analysis to real-world wheels.
+
+\section*{ACKNOWLEDGEMENTS}
+This material is based upon work supported by the National Science Foundation Graduate Research Fellowship Program under Grant No.
+DGE-1324585.
+Any opinions, findings, and conclusions or recommendations expressed in this material are those of the authors and do not necessarily reflect the views of the National Science Foundation.
 
@@
 $\bar{\bm{k}}^{el},\bar{\bm{k}}^{tens}$ & Part of $\bar{\bm{k}}$ proportional to $E_sA_s$ and $T$, respectively\\
-$\bar{k}_{ij}$ & Tensor components of $\bar{\bm{k}}$\\
+$\bar{k}_{ij}$ & Matrix components of $\bar{\bm{k}}$\\
 $\e,\ee,\eee$ & Coordinate axes defined at the rim shear center\\
@@
 			  equal to the buckling load for a straight beam-column of length $2\pi R$\\
-$R$			& Rim radius--measured from the rim shear center\\
-$E$			& Rim Young's modulus\\
+$R$			& Rim radius measured from the rim shear center\\
+$E, G$		& Rim Young's modulus and shear modulus\\
 $I_{11}$	& Second moment of area of the rim in the lateral direction\\

### 2016-08-31 latex/paper.tex
minor edits to v1

@@
 \section*{ABSTRACT}
-Bicycle spokes brace and stiffen the slender rim, but only as long as they do not lose pre-tension under external loads.
-However, above a certain critical tension, the rim will buckle into a non-planar saddle shape.
+The spokes of a bicycle wheel brace and stiffen its slender rim, but only as long as the pre-tension is high enough to prevent them from going slack under external loads.
+However, above a certain critical spoke tension, the rim will buckle into a non-planar saddle shape.
 Despite its importance for wheel strength and safety, there is no formula in the literature for the maximum tension a wheel can withstand before buckling.
@@
 Here we derive such a formula.
-The non-dimensionalized buckling tension depends only on two parameters which are identified as the ratio of torsional stiffness to bending stiffness, and the ratio of lateral spoke bracing stiffness to rim bending stiffness.
+The non-dimensionalized buckling tension depends only on two parameters which are identified as the ratio of rim torsional stiffness to rim bending stiffness, and the ratio of lateral spoke bracing stiffness to rim bending stiffness.
 The precise buckled shape is determined by these two stiffness ratios and the number of positive peaks of the lateral displacement may in general by greater than 2.
@@
 The earliest bicycles were built in the workshops of carriage makers who adapted the wheel technology of the time.
-These wheels typically had iron tires, wooden rims and stout wooden spokes which bore the weight of the machine and rider in compression~\cite{Herlihy2004a}.
-The spokes needed to be wide enough to prevent lateral buckling and to keep the rim planar, which made them very heavy and limited their diameter.
+These wheels typically had iron tires, wooden rims, and stout wooden spokes which bore the weight of the machine and rider in compression~\cite{Herlihy2004a}.
+The spokes needed to be wide enough to prevent lateral buckling and to keep the rim planar, which made them very heavy and limited their maximum diameter.
 Around 1869, the tension-spoke wheel, an idea proposed in 1808 by Sir George Cayley but never realized, found its first application on the rapidly-developing bicycle~\cite{Ackroyd2011a,Clayton1991b}.
@@
 The tension-spoke wheel is a prestressed structure and is susceptible to buckling under internal forces.
-As the spokes are tightened uniformly, the rim deforms radially under uniform compression.
-At a critical tension, the rim buckles out of its initial plane.
-The post-buckling load-displacement curve is generally stable and the original shape of the wheel can often be recovered by reducing the tension.
+As the spokes are tightened uniformly, the rim shrinks radially under compression.
+At a critical spoke tension, the rim becomes unstable and buckles into a non-planar shape.
+The post-buckling tension vs.
+displacement curve is generally stable and the original shape of the wheel can often be recovered by reducing the tension.
 Buckling can also be triggered by external forces in an otherwise stable wheel leading to a release of strain energy.
@@
 Despite its implications for wheel strength, the buckling problem has never received a rigorous treatment to our knowledge.
-Jobst Brandt alludes to buckling in his practical manual for wheelbuilders~\cite{Brandt1993c}:
+Jobst Brandt alludes to buckling due to uniform tension in his practical manual for wheelbuilders~\cite{Brandt1993c}:
 
@@
 He did not discuss the problem of buckling under tension further, but proposed that wheel failure commonly occurs due to lateral buckling when external loads have caused spokes to lose tension.
-Pippard and Francis~\cite{Pippard1932d} derived an equation for lateral deflection based on an elastic foundation model, but did not discuss stability and neglected any effects of spoke tension.
+Pippard and Francis~\cite{Pippard1932d} derived an equation for lateral deflection based on an elastic foundation model, but did not discuss stability and neglected the effect of spoke tension.
 
@@
 The spokes are kept under tension by the rim.
-Without pretension, the spokes would immediately buckle under a very small load due to their high aspect ratio.
+Without pretension, the spokes would immediately buckle under a very small axial load due to their high aspect ratio.
 
@@
 	\end{equation}
-The warping energy is primarily related to the bending energy in the sidewalls (which act as beam flanges) of the rim\footnote{See http://www.steel-insdag.org/TeachingMaterial/Chapter17.pdf for a simple treatment.}.
-The warping constant $I_w$ has units of m$^6$, and for single-wall rims the warping stiffness $EI_w$ generally exceeds the torsion stiffness $GJ$.
+The warping energy is primarily related to the bending energy in the rim sidewalls (which act as I-beam flanges) of the rim due to twisting\footnote{See http://www.steel-insdag.org/TeachingMaterial/Chapter17.pdf for a simple treatment.}.
+For a single-wall rim, the energy contribution from warping can exceed the contribution from torsion.
 
@@
 The tension stiffness is the component responsible for the transverse vibrations of a guitar string, e.g.
-The term $\n\n$ is the dyadic product, or tensor product (also called the outer product), and is defined by the property $\n\n\cdot \bm{v} = (\n\cdot \bm{v})\n$ for any vector $\bm{v}$.
+The term $\n\n$ is the dyadic product, or tensor product (also called the outer product), and is defined by the property $\n\n\cdot \bm{a} = (\n\cdot \bm{a})\n$ for any vector $\bm{a}$.
 
@@
 The continuum spoke stiffness can be written $\bar{\bm{k}} = \bar{\bm{k}}^{el} + \bar{\bm{k}}^{tens}$, where the second term is proportional to $T$.
-A very close approximation for the $\bar{k}_{uu}$ component for a symmetrically dished wheel is given by
+A very close approximation for the lateral component $\bar{k}_{uu}$ for a symmetrically dished wheel is given by
 	\begin{equation}\label{eq:kuu}
@@
 \ref{fig:Tc_nondim} (a)).
-However, two important \textit{approximate} solutions can be derived by replacing the discrete $n$ with a continuous variable $\bar{n}$ and minimizing Eqn.
+However, two important \textit{approximate} solutions can be derived by replacing the discrete mode number $n$ with a continuous variable $\bar{n}$ and minimizing Eqn.
 \ref{eq:Tc_nondim} analytically, resulting in two simple power laws for $T_c$.
@@
 	\end{equation}
-Following the same procedure in section \ref{sec:powerlaw_1} gives a separate power law, after simplification.
+Following the same procedure in section \ref{sec:powerlaw_1} gives another separate power law:
 	\begin{equation}\label{eq:Tc_boef}
@@
 	\begin{equation}\label{eq:Kbending}
-	K_{torsion} = \frac{\pi EI_{11}}{R^3}(n^2-1)^2
+	K_{bending} = \frac{\pi EI_{11}}{R^3}(n^2-1)^2
 	\end{equation}
@@
 \section{CONCLUSIONS}
-The bicycle wheel relies on spoke tension to support large loads in the radial direction.
+The bicycle wheel relies on spoke tension to support external loads.
 As a rule-of-thumb, a typical wheel can withstand a radial load of approximately $2T$ before the bottom-most spokes lose tension.
@@
 Therefore they should be in relative balance.
-Modern double-walled rims are advantageous in this respect because of their closed cross-section, while classic single-walled rims are much more flexible in torsion.
+Modern double-walled rims are advantageous in this respect because of their closed cross-section, while classic single-walled rims are much more flexible in torsion than in bending.
 \textbf{(2)} Since the rim and spoke stiffnesses act like springs in parallel, they can both be increased to roughly equal effect.
-\textbf{(3)} Since the lateral spoke stiffness scales roughly as $\sin^2{\alpha}$, the most significant factor affecting the spoke stiffness is the cone angle $\alpha$.
+\textbf{(3)} Since the lateral spoke stiffness scales as $\sin^2{\alpha}$, the most significant factor affecting the spoke stiffness is the cone angle $\alpha$.
 Even a modest increase in hub width can increase the stiffness and buckling tension dramatically, while having negligible effect on the radial stiffness.

### 2016-08-31 latex/paper.tex
Manuscript v2

@@
 The spokes of a bicycle wheel brace and stiffen its slender rim, but only as long as the pre-tension is high enough to prevent them from going slack under external loads.
-However, above a certain critical spoke tension, the rim will buckle into a non-planar saddle shape.
+However, above a certain critical spoke tension, the rim will buckle into a non-planar shape.
 Despite its importance for wheel strength and safety, there is no formula in the literature for the maximum tension a wheel can withstand before buckling.
@@
 The non-dimensionalized buckling tension depends only on two parameters which are identified as the ratio of rim torsional stiffness to rim bending stiffness, and the ratio of lateral spoke bracing stiffness to rim bending stiffness.
-The precise buckled shape is determined by these two stiffness ratios and the number of positive peaks of the lateral displacement may in general by greater than 2.
+The buckled rim adopts a sinusoidal lateral profile where the number of positive peaks $n\geq 2$ is determined by these two stiffness ratios.
 We show that the wheel may be represented by a system of equivalent springs: two series-connected springs for the rim bending and torsion stiffness, connected in parallel with a spring for the spoke system.

### 2016-09-01 latex/paper.tex
Minor edits

@@
 The earliest bicycles were built in the workshops of carriage makers who adapted the wheel technology of the time.
-These wheels typically had iron tires, wooden rims, and stout wooden spokes which bore the weight of the machine and rider in compression~\cite{Herlihy2004a}.
+These wheels typically had iron tires, wooden rims, and stout wooden spokes which bore the weight of the machine and rider in compression~\cite{Herlihy2004}.
 The spokes needed to be wide enough to prevent lateral buckling and to keep the rim planar, which made them very heavy and limited their maximum diameter.
@@
 At a critical spoke tension, the rim becomes unstable and buckles into a non-planar shape.
-The post-buckling tension vs.
-displacement curve is generally stable and the original shape of the wheel can often be recovered by reducing the tension.
+The average spoke tension remains roughly constant after buckling and the original shape of the wheel can often be recovered by loosening the spokes.
 Buckling can also be triggered by external forces in an otherwise stable wheel leading to a release of strain energy.

### 2016-09-01 latex/paper.tex
Started FEA results table

@@
 \section*{LIST OF TERMS}
-A partial list of symbols is given below. Other symbols are defined in-place on not re-used.
+A partial list of symbols is given below. Other symbols are defined in-place or not re-used.
 
@@
 
+\newpage
+\section*{FINITE-ELEMENT RESULTS}
+For all the following wheels: $R=0.3$ m, $EI_{11}=EI_{22}=69.0$ N-m$^2$, $GJ=26.0$ N-m$^2$,
+
+\begin{tabular}{c|c|c}
+$n_s$ & Hub flange width [mm] & Hub flange diameter [mm]\\
+\hline
+
+\end{tabular}
+
 \end{document}

### 2016-09-01 latex/paper.tex
Complete FEA table

--- 2016-09-01 latex/paper.tex
@@
 \section*{LIST OF TERMS}
-A partial list of symbols is given below. Other symbols are defined in-place on not re-used.
 
+\begin{table}[ht!]
+\caption{A partial list of symbols is given below. Other symbols are defined in-place or not re-used.}
 \begin{tabular}{c|p{12cm}}
@@
 \end{tabular}
+\end{table}
+
+\newpage
+\section*{FINITE-ELEMENT RESULTS}
+For all the following wheels: $R=0.3$ m, $EI_{11}=EI_{22}=69.0$ N-m$^2$, $GJ=26.0$ N-m$^2$, $I_w=0$, $E_s=210$ GPa.
+The cross-sectional area of the rim is 100 mm$^2$.
+All length measurements in the following table are given in mm.
+$T_c$ is reported in Newtons.
+Hub flange width is the distance from the drive-side flange to the left-side flange.
+Hub flange diameter is the diameter of the circle that passes through the centers of the spoke eyelets.
+
+\begin{table}[ht!]
+\caption{Parameters for FEA simulations shown in Fig \ref{fig:Tc_nondim} (b).}
+\begin{tabular}{c|c|c|c|c|c}
+\hline
+$n_s$	& Spoke diameter	& Hub flange width	& Hub flange diameter	& $T_c$ ABAQUS	& $T_c$ theory\\
+\hline
+36		& 0.693				& 50				& 40					& 477			& 475\\
+36		& 0.981				& 50				& 40					& 668			& 704\\
+\hline
+48		& 1.040				& 50				& 40					& 694			& 698\\
+48		& 1.201				& 50				& 40					& 862			& 870\\
+48		& 1.343				& 50				& 40					& 956			& 942\\
+48		& 1.471				& 50				& 40					& 1021			& 1005\\
+48		& 1.900				& 50				& 40					& 1270			& 1259\\
+48		& 1.943				& 60				& 40					& 1563			& 1573\\
+\hline
+64		& 1.671				& 60				& 40					& 1450			& 1418\\
+\end{tabular}
+\end{table}
+
+For all the following wheels: $n_s=36$, $R=0.3$ m, hub flange width = $50$ mm, hub flange diameter = $40$ mm, spoke diameter = $0.981$ mm, $EI_{11}=EI_{22}=69.0$ N-m$^2$, $I_w=0$, $E_s=210$ GPa.
+$GJ$ varies with $\mu$ given in the table below.
+The cross-sectional area of the rim is 100 mm$^2$.
+$T_c$ is reported in Newtons.
+
+\begin{table}[ht!]
+\centering
+\caption{Parameters for FEA simulations shown in Fig \ref{fig:Tc_nondim} (c).}
+\begin{tabular}{c|c|c}
+\hline
+$\mu$	& $T_c$ ABAQUS & $T_c$ theory\\
+\hline
+0.001	& 154	& 121 \\
+0.00215	& 191	& 157 \\
+0.00464	& 220	& 212 \\
+0.01	& 264	& 258 \\
+0.0215	& 359	& 344 \\
+0.0464	& 501	& 487 \\
+0.1		& 556	& 574 \\
+0.215	& 624	& 646 \\
+0.464	& 701	& 723 \\
+1		& 759	& 785 \\
+\end{tabular}
+\end{table}
 

### 2016-09-01 latex/paper.tex
Remove underlines from references
Minor edits

@@
 The spokes of a bicycle wheel brace and stiffen its slender rim, but only as long as the pre-tension is high enough to prevent them from going slack under external loads.
-However, above a certain critical spoke tension, the rim will buckle into a non-planar shape.
-Despite its importance for wheel strength and safety, there is no formula in the literature for the maximum tension a wheel can withstand before buckling.
+However, if the spoke tension is too high, the rim will buckle into a non-planar shape.
+Despite its importance for wheel strength and safety, we know of no formula in the literature for the maximum tension a wheel can withstand before buckling.
 
@@
 The buckled rim adopts a sinusoidal lateral profile where the number of positive peaks $n\geq 2$ is determined by these two stiffness ratios.
-We show that the wheel may be represented by a system of equivalent springs: two series-connected springs for the rim bending and torsion stiffness, connected in parallel with a spring for the spoke system.
+The lateral stiffness of the wheel may be represented by a system of equivalent springs: two series-connected springs for the rim bending and torsion stiffness, connected in parallel with a spring for the spoke system.
 
@@
 He did not discuss the problem of buckling under tension further, but proposed that wheel failure commonly occurs due to lateral buckling when external loads have caused spokes to lose tension.
-Pippard and Francis~\cite{Pippard1932d} derived an equation for lateral deflection based on an elastic foundation model, but did not discuss stability and neglected the effect of spoke tension.
+Pippard and Francis~\cite{Pippard1932d} derived an equation for lateral deflection based on an elastic foundation model, but did not discuss stability and neglected the effect of spoke pre-tension.
 
-The theory of bicycle wheel buckling is closely related to the theory of flexural-torsional buckling of arches.
+The theory of bicycle wheel buckling is closely related to the theory of flexural-torsional buckling of rings and arches.
 Timoshenko and Gere~\cite{Timoshenko1961a} gave a formula for the critical load for a ring with doubly-symmetric cross-section subjected to radial loads.
@@
 Large observation wheels such as the London Eye~\cite{Mann2001a} and the Singapore Flyer~\cite{Allsop2009a} resemble large bicycle wheels with lateral stability due to the bracing angle of prestressed cables, and must be designed against flexural-torsional buckling.
+
+The buckling tension is controlled by the geometric and material properties of the wheel.
+The wheel designer may wish to achieve the stiffest and strongest wheel for a given size and weight by tuning parameters such as spoke count, spoke angle, rim width and depth, and material.
+Here we will answer the question of precisely how each of these factors affects the maximum safe tension in order to guide rational wheel design and component selection.
 
@@
 \ref{fig:schematic}.
-Increasing the number of spokes only improves the estimate.} was given by Sharp~\cite{Sharp1977a} by considering a force balance on the top (or bottom) half of the rim.
+Increasing the number of spokes only improves the estimate.} of the compressive force was given by Sharp~\cite{Sharp1977a} by considering a force balance on the top (or bottom) half of the rim.
 	\begin{equation}\label{eq:TN}
@@
 	\end{equation}
-The warping energy is primarily related to the bending energy in the rim sidewalls (which act as I-beam flanges) of the rim due to twisting\footnote{See http://www.steel-insdag.org/TeachingMaterial/Chapter17.pdf for a simple treatment.}.
+The warping energy $EI_w(\kappa_3')^2$ is primarily related to the bending energy in the rim sidewalls (which act as I-beam flanges) of the rim due to twisting\footnote{See http://www.steel-insdag.org/TeachingMaterial/Chapter17.pdf for a simple treatment.}.
 For a single-wall rim, the energy contribution from warping can exceed the contribution from torsion.
@@
 	\end{equation}
-since \n{}, \nn{}, \nnn{} are mutually orthogonal, the total force on the rim is given by $\bm{f} = \bm{k}_f \cdot \bm{u}_n$, where the spoke stiffness matrix is given by
+Since \n{}, \nn{}, \nnn{} are mutually orthogonal, the total force on the rim is given by $\bm{f} = \bm{k}_f \cdot \bm{u}_n$, where the spoke stiffness matrix is given by
 	\begin{equation}\label{eq:kf}
@@
 	\begin{split}
-	u &= \delta u_n \cos{\frac{ns}{R}} \\
-	\phi &= \delta\phi_n \cos{\frac{ns}{R}}
+	u &= \delta u_n \cos{\left(\frac{ns}{R}\right)}\\
+	\phi &= \delta\phi_n \cos{\left(\frac{ns}{R} \right)}
 	\end{split}
@@
 \,\,\,\, where the ``Euler tension'' $T_e=2\pi EI_{11}/n_sR^2$ is the tension that would produce a compressive force in the rim equal to the classical buckling load for a straight beam-column of length $2\pi R$.
-The parameter $\tilde{\mu} = \widetilde{GJ}/EI_{11}$ is the ratio of the effective torsional stiffness to the lateral bending stiffness.
+The parameter $\tilde{\mu} = \widetilde{GJ}/EI_{11}$ is the ratio of the effective torsional stiffness to lateral bending stiffness.
 The parameter $\lambda = \bar{k}_{uu}^{el}R^4/EI_{11}$ is the ratio of the spoke system stiffness $\bar{k}_{uu}^{el}R$ to the rim bending stiffness $EI_{11}/R^3$.
@@
 \ref{eq:Tc_nondim} and finite-element calculations.
-The power law is strictly less than Eqn.
-\ref{eq:Tc_nondim} which conveniently gives a conservative design criterion.
+For $\tilde{\mu}<0.13$, the power law is strictly less than Eqn.
+\ref{eq:Tc_nondim} which gives a conservative design criterion.
 
@@
 	\end{equation}
-Following the same procedure in section \ref{sec:powerlaw_1} gives another separate power law:
+Following the same procedure in section \ref{sec:powerlaw_1} gives another power law:
 	\begin{equation}\label{eq:Tc_boef}
@@
 	\end{equation}
-The form of Equation \ref{eq:Kn_spring} suggests an equivalent spring model (Fig.
-\ref{fig:Kn}) for the $n$th mode in which the rim stiffness is represented by two springs connected which is then connected in parallel with the spoke stiffness.
-If there is a large difference between the rim bending stiffness and torsion stiffness, the equivalent rim stiffness will be dominated by the smaller of the two.
+Equation \ref{eq:Kn_spring} corresponds to the equivalent spring model (Fig.
+\ref{fig:Kn}) for the $n$th mode.
+Since the rim bending stiffness are connected in series, the equivalent rim stiffness will be dominated by the smaller of the two.
 
@@
 \textbf{(2)} Since the rim and spoke stiffnesses act like springs in parallel, they can both be increased to roughly equal effect.
-\textbf{(3)} Since the lateral spoke stiffness scales as $\sin^2{\alpha}$, the most significant factor affecting the spoke stiffness is the cone angle $\alpha$.
+\textbf{(3)} Since the lateral spoke stiffness scales as $\sin^2{\alpha}$, the most significant factor affecting the lateral spoke system stiffness is the cone angle $\alpha$.
 Even a modest increase in hub width can increase the stiffness and buckling tension dramatically, while having negligible effect on the radial stiffness.
@@
 The spoke system lateral stiffness (Eqn.
-\ref{eq:kuu}) scales directly with the elastic modulus (material-dependent, but not alloy-dependent) and the cross-sectional area, and inversely with the spoke length (which varies only slightly depending on spoke pattern).
+\ref{eq:kuu}) scales directly with the elastic modulus and the cross-sectional area, and inversely with the spoke length (which varies only slightly depending on spoke pattern).
 It is counter-intuitive that increasing the number of spokes actually slightly \textit{decreases} the buckling tension (due to the factor of $n_s$ appearing in $T_e$ in Eqn.
@@
 However, increasing the number of spokes also reduces the fraction of load transferred to a single spoke and thus reduces the required tension.
-The net effect is that more spokes results in a stronger wheel.
+The net effect is that more spokes generally results in a stronger wheel.
 
@@
 
+\normalem
 \bibliographystyle{naturemag}
@@
 $\bar{\bm{k}}$ & Equivalent stiffness per-unit-length of the spoke system. $\bar{\bm{k}} = \bar{\bm{k}}^{el} + \bar{\bm{k}}^{tens}$\\
-$\bar{\bm{k}}^{el},\bar{\bm{k}}^{tens}$ & Part of $\bar{\bm{k}}$ proportional to $E_sA_s$ and $T$, respectively\\
-$\bar{k}_{ij}$ & Matrix components of $\bar{\bm{k}}$\\
+$\bar{\bm{k}}^{el},\bar{\bm{k}}^{tens}$ & Parts of $\bar{\bm{k}}$ proportional to $E_sA_s$ and $T$, respectively\\
+$\bar{k}_{ij}$ & Matrix components of $\bar{\bm{k}}$, where $i,j\rightarrow u,w,v$ \\
 $\e,\ee,\eee$ & Coordinate axes defined at the rim shear center\\
@@
 $\tilde{\mu}$ & Ratio of effective torsional rigidity to bending rigidity: $\widetilde{GJ}/EI_{11}$\\
-$\lambda$	& Ratio of spoke system lateral stiffness to rim bending stiffness OR ratio
-			  of rim radius to the characteristic length of a beam on an elastic foundation.\\
+$\lambda$	& Ratio of spoke system lateral stiffness to rim effective bending stiffness \\
 \hline
@@
 64		& 1.671				& 60				& 40					& 1450			& 1418\\
+\hline
 \end{tabular}
@@
 1		& 759	& 785 \\
+\hline
 \end{tabular}

### 2016-09-01 latex/paper.tex
Rewrite final Intro paragraph

@@
 
-The buckling tension is controlled by the geometric and material properties of the wheel.
-The wheel designer may wish to achieve the stiffest and strongest wheel for a given size and weight by tuning parameters such as spoke count, spoke angle, rim width and depth, and material.
-Here we will answer the question of precisely how each of these factors affects the maximum safe tension in order to guide rational wheel design and component selection.
+The spoke tension to cause buckling is affected by the spoke axial stiffness, spoke count, and spoke bracing angle.
+It is also strongly affected by the rim lateral bending stiffness, torsional stiffness, and radius.
+Our aim here is to provide formulas to predict the maximum allowable tension in terms of these properties and offer practical design rules to select the properties to achieve a desired maximum tension.
 
@@
 \begin{table}[ht!]
-\caption{A partial list of symbols is given below. Other symbols are defined in-place or not re-used.}
+\caption{A partial list of symbols used in this paper is given below. Other symbols are defined in-place or not re-used.}
 \begin{tabular}{c|p{12cm}}
